---
title: "Preparation of Heart Disease Dataset"
format: html
---

This file includes all steps that we took to prepare the original data for the analyses that we conducted in our study.

<!-- [This is a comment, it won't be shown as text if you render this file.] 

     If your original data is >100Mb, you cannot include it in your GitHub repo.
     Make sure every team member has the dataset stored on their computer, 
     in this project folder, at the same path, so you can load it with the same 
     code. But don't push it. You can add the path to the dataset to the file 
     .gitignore, and this way you will never be prompted to commit it.
     
     If your GitHub repo does not have your original data file, add a statement 
     below this comment that explains this and provides a link to a Google drive 
     folder. You can copy the statement you wrote in Homework 7.
-->

## Load Libraries

These are the R libraries needed for the preparation steps:

```{r}
#| message: FALSE
library(tidyverse)
library(janitor)
library(tidyr)
library(readr)
```

## Load Original Data

Here we load the original dataset used in our study:

```{r}
#| message: FALSE
heart_data <- read_csv("disease-raw.csv")
```

Here we clean the variable names using the function `clean_names()`:

```{r}
heart_data <- clean_names(heart_data)
```


## Prepare Data

### Numeric Variables

[Summary of Numeric Variables]{.underline}

```{r}
#| message: FALSE
# numeric variables: bmi, physical_health, mental_health, sleep_time

summary(heart_data)
```

The numeric variables of this dataset include the following: `bmi`, `physical_health`, `mental_health`, and `sleep_time`.

`bmi` contains enexpected values, including a minimum = 12.02. A BMI of 12 is not normal and indicates being severely underweight. Such values will be treated as NA. `sleep_time` also contains unexpected values (in hours), including a minimum = 1 and a maximum = 24. Although plausible, 1 or 24 hours of sleep is extremely concerning and may influence data analyses. These will also be treated as NA.

[Mutating Variables]{.underline}

```{r}
#| message: FALSE

# mutating bmi
heart_data <- heart_data %>% 
  mutate(
    bmi = case_when(
      bmi < 15 ~ NA,
      bmi > 60 ~ NA,
      TRUE ~ bmi
    )
  )

# mutating sleep_time

heart_data <- heart_data %>% 
  mutate(
    sleep_time = case_when(
      sleep_time < 3 ~ NA,
      sleep_time > 16 ~ NA,
      TRUE ~ sleep_time
    )
  )
```

Here, implausible values in `bmi` and `sleep_time` have been recoded as NA. Specifically, BMI values belo 15 or above 60 are treated as unrealistic, while sleep times below 3 hours or above 16 hours are flagged as extreme and set to NA. All other values are retained as originally recorded. This ensures that outliers and likely data entry errors do not distort subsequent analyses.

### Binary and Categorical Variables

[Summary of Binary and Categorical Variables]{.underline}

```{r}
#| message: FALSE

summary(heart_data)

# binary variables: heart_disease, smoking, alcohol_drinking, stroke, diff_walking, physical_activity, asthma, kidney_disease, skin_cancer, sex

# other categorical variables: age_category, race, diabetic, gen_health
```

None of the binary and other categorical variables contain unexpected or NA values.

[Recoding Binary Variables into a Numeric Format]{.underline}

```{r}
# changing levels of binary variables: heart_disease, smoking, alcohol_drinking, stroke, diff_walking, physical_activity, asthma, kidney_disease, skin_cancer, sex

heart_data <- heart_data %>% 
  mutate(
    heart_disease = ifelse(heart_disease == "Yes", 1, 0),
    smoking = ifelse(smoking == "Yes", 1, 0),
    alcohol_drinking = ifelse(alcohol_drinking == "Yes", 1, 0),
    stroke = ifelse(stroke == "Yes", 1, 0),
    diff_walking = ifelse(diff_walking == "Yes", 1, 0),
    physical_activity = ifelse(physical_activity == "Yes", 1, 0),
    asthma = ifelse(asthma == "Yes", 1, 0),
    kidney_disease = ifelse(kidney_disease == "Yes", 1, 0),
    skin_cancer = ifelse(skin_cancer == "Yes", 1, 0),
    sex = ifelse(sex == "Female", 1, 0)
  )
```

We recorded several binary variables into numeric form. "Yes" responses (and "Female" for sex) are converted to 1, while "No" responses (and "Male" for sex) are converted to 0. This allows the variables to be used more effectively in analysis and modeling.

[Recoding `Diabetic` Variable into a Numeric Format]{.underline}

```{r}
# changing levels of diabetic variable

table(heart_data$diabetic) 
# No | No, borderline diabetes | Yes | Yes (during pregnancy)

heart_data <- heart_data %>% 
  mutate(
    diabetic = case_when(
      diabetic %in% c("No", "No, borderline diabetes") ~ "No",
      diabetic %in% c("Yes", "Yes (during pregnancy)") ~ "Yes"
    )
  )

heart_data <- heart_data %>% 
  mutate(
    diabetic = ifelse(diabetic == "Yes", 1, 0)
    )
```

We simplified the `diabetic` variable by combining related categories into two groups: "No" (including borderline diabetes) and "Yes" (including diabetes during pregnancy). We then recoded the variable into numeric form, where "Yes" is represented as 1 and "No" as 0. This creates a clear binary variable that is easier to use in analysis and modeling.

[Mutating Binary and Categorical Variables]{.underline}

```{r}
heart_data <- heart_data %>% 
  mutate(
    heart_disease = as_factor(heart_disease),
    smoking = as_factor(smoking),
    alcohol_drinking = as_factor(alcohol_drinking),
    stroke = as_factor(stroke),
    diff_walking = as_factor(diff_walking),
    physical_activity = as_factor(physical_activity),
    asthma = as_factor(asthma),
    kidney_disease = as_factor(kidney_disease),
    skin_cancer = as_factor(skin_cancer),
    sex = as_factor(sex)
  )

# changing age_category
heart_data <- heart_data %>% 
  mutate(age_category = case_when(
    age_category %in% c("18-24", "25-29", "30-34") ~ "18-34",
    age_category %in% c("35-39", "40-44", "45-49", "50-54") ~ "35-54",
    age_category %in% c("55-59", "60-64", "65-69", "70-74", "75-79", "80 or older") ~ "55 and older"
  ))

heart_data <- heart_data %>% 
  mutate(age_category = case_when(
    age_category == "18-34" ~ 1,
    age_category == "35-54" ~ 2,
    age_category == "55 and older" ~ 3
  ))

heart_data <- heart_data %>%
  mutate(
    age_category = as_factor(age_category)
  )
```

```{r}
# ordering variables

heartdisease$gen_health <- factor(
  heartdisease$gen_health, 
  levels = c("Poor", "Fair", "Good", "Very Good", "Excellent")
```

## Save Data

In the following chunk, we save our dataset as an RDS file using the function `saveRDS()`. 

```{r}
saveRDS(heart_data, "heart-prepared.RDS")
```






