---
title: "Preparation of Heart Disease dataset"
format: html
---

This file includes all steps that we took to prepare the original data for the analyses that we conducted in our study.

<!-- [This is a comment, it won't be shown as text if you render this file.] 

     If your original data is >100Mb, you cannot include it in your GitHub repo.
     Make sure every team member has the dataset stored on their computer, 
     in this project folder, at the same path, so you can load it with the same 
     code. But don't push it. You can add the path to the dataset to the file 
     .gitignore, and this way you will never be prompted to commit it.
     
     If your GitHub repo does not have your original data file, add a statement 
     below this comment that explains this and provides a link to a Google drive 
     folder. You can copy the statement you wrote in Homework 7.
-->

## Load Libraries

These are the R libraries needed for the preparation steps:

```{r}
#| message: FALSE
library(tidyverse)
library(janitor)
library(tidyr)
library(readr)
```

## Load Original Data

Here we load the original dataset used in our study:

```{r}
#| message: FALSE
heart_data <- read_csv("disease-raw.csv")
```

Here we clean the variable names using the function `clean_names()`:

```{r}
heart_data <- clean_names(heart_data)
```


## Prepare Data

### Prepare Numeric Variables

```{r}
#| message: FALSE
# numeric variables: bmi, physical_health, mental_health, sleep_time

summary(heart_data)
```

The numeric variables of this dataset include the following: `bmi`, `physical_health`, `mental_health`, and `sleep_time`.

`bmi` contains enexpected values, including a minimum = 12.02. A BMI of 12 is not normal and indicates being severely underweight. Such values will be treated as NA. `sleep_time` also contains unexpected values (in hours), including a minimum = 1 and a maximum = 24. Although plausible, 1 or 24 hours of sleep is extremely concerning and may influence data analyses. These will also be treated as NA.

<!-- [This is a comment, it won't be shown as text if you render this file.]

     This section is unstructured because the sequence of preparation steps
     that you take here are specific to your particular datasets and research 
     questions. For example, you may or may not need to join multiple datasets,
     pivot the data, filter observations, mutate variables, etc. The sequence 
     in which you need to do these operations (e.g. join and then add variables, 
     or viceversa) is also specific to your case.
     
     You should include in this section all and only the preparation steps that  
     are necessary (leave out any unnecessary steps that you did in HW 7 for
     practice). 
     
     Break down the steps into subtasks, each executed in a separate chunk and 
     preceded by a short explanation of what the chunk will achieve. 
     For example:
     "Prepare the datasets to be joined" followed by a chunk where you change
     some of the country names so all countries have the same name in both
     datasets and R can join them. 
     "Join datasets X and Z" followed by a chunk where you join the datasets.
     "Filter observations from 2010" followed by a chunk where you filter.
     "Remove unnecessary variables" followed by a chunk where you select away 
     some variables.
     Etc. etc.
-->

## Save Data

In the following chunk, we save our dataset as an RDS file using the function `saveRDS()`. 

```{r}

```






