---
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| include: false

library(tidyverse)
library(tidyr)
library(kableExtra)
```

## 2. Exploratory Data Analysis & Analysis Approach

**Data Set 1: Diabetic Patient Hospital Readmission in the US**

***Response Variable: Hospital Readmission***

```{r}
#| include: false
# PREPPING DATA & VARIABLES
diabetes_data <- readRDS("../diabetes_data/diabetes_data-prepared.RDS")

# give meaningful label to the readmitted binary variable
diabetes_data$readmitted_binary <- factor(
  diabetes_data$readmitted_binary,
  levels = c(0, 1),
  labels = c("Late Readmission (>30 Days)",
             "Early Readmission (<30 Days)")
)

# assign counts to variables
less_than_thirty<- diabetes_data |>
  filter(readmitted == "<30") |>
  count()
less_than_thirty_prop<- round(less_than_thirty/nrow(diabetes_data)*100, 2)

greater_than_thirty <- diabetes_data |>
  filter(readmitted == ">30") |>
  count()
greater_than_thirty_prop <- round(greater_than_thirty/nrow(diabetes_data)*100, 2)

no_readmission <- diabetes_data |>
  filter(readmitted == "NO") |>
  count()
no_readmission_prop <- round(no_readmission/nrow(diabetes_data)*100, 2)

late_readmission <- greater_than_thirty + no_readmission
late_readmission_prop <- round(late_readmission/nrow(diabetes_data)*100,2)

```

We will take the hospital inpatient readmission indicator (`readmitted`)
as our primary predictory variable. In our cleaned dataset, this binary
variable indicates whether a diabetic patient was a) readmitted within
30 days of their first admission, i.e. early readmission or b)
readmitted \>30 days after their first admission or not at all, i.e.
late readmission.

The vast majority of diabetic patients (n = `r late_readmission`;
`r late_readmission_prop`% of all patients) were readmitted only after a
month (n = `r greater_than_thirty`;`r greater_than_thirty_prop`% of all
patients) or were never readmitted (n = `r no_readmission`;
`r no_readmission_prop`% of all patients). The remaining
`r less_than_thirty` (`r less_than_thirty`%) patients were readmitted
within 30 days of their first admission (@fig-readmission-freq;
@tbl-readmission-freq).

::: {.layout-ncol=2}
```{r}
#| include: false

# readmitted - NONBINARY
ggplot(data = diabetes_data, aes(x = factor(readmitted), fill = readmitted)) +
  geom_bar(color = "black") + 
  labs(x = "Hospital Readmission Status",
       y = "Count") +
  scale_fill_manual(values = c("<30" = "tomato", ">30" = "skyblue", "NO" = "steelblue")) +
  theme_classic()

```

```{r}
#| echo: false
#| label: fig-readmission-freq
# readmitted - BINARY
ggplot(data = diabetes_data, aes(x = factor(readmitted_binary), fill = readmitted)) +
  geom_bar(color = "black") + 
  labs(x = "Hospital Readmission Status",
       y = "Count") +
  #scale_x_discrete(labels = c("0" = "Late Readmission (>30 Days)",
  #                            "1" = "Early Readmission (<30 Days)")) +
  scale_fill_manual(values = c("<30" = "tomato", ">30" = "skyblue", "NO" = "steelblue")) +
  theme_classic()
```

```{r}
#| echo: false
#| label: tbl-readmission-freq
#| tbl-cap: "Hospital Readmission Frequency for Diabetic Patients in the US"

# table of readmission counts - NONBINARY
diabetes_data |> 
  count(readmitted) |>
  mutate(percentage = (n/nrow(diabetes_data))*100) |>
  mutate(percentage = round(percentage, 2)) |>
  kable(col.names = c("Readmission Status", "Count", "Percentage (%)"))

# table of readmission counts - BINARY
diabetes_data |> 
  count(readmitted_binary) |>
  mutate(percentage = (n/nrow(diabetes_data))*100) |>
  mutate(percentage = round(percentage, 2)) |>
  kable(col.names = c("Readmission Status", "Count", "Percentage (%)"))
```
:::

***Key Predictors: HbA1c test, number of diagnoses, serum glucose, drug
tests, etc.***

\- **1-3 important predictors**: - description - visualizations -
summary statistics

```{r}
# give meaningful label to the HbA1c binary variable
diabetes_data$a1cresult_binary <- factor(
  diabetes_data$a1cresult_binary,
  levels = c(0, 1),
  labels = c("No Test Administered",
             "Test Administered")
)
```

```{r}
#| echo: false
#| label: fig-readmission-hba1c

table(diabetes_data$a1cresult)

ggplot(data = diabetes_data, aes(x = factor(a1cresult_binary), fill = a1cresult)) +
  geom_bar(color = "black") + 
  labs(x = "Hemoglobin A1c Test Result",
       y = "Count") +
  scale_fill_manual(values = c(">7" = "orange", ">8" = "red",  "Norm" = "springgreen2", "None" = "gray")) +
  theme_classic()
```

***Other Predictors***

\- **other available predictors** (list them in a paragraph & provide
range of values, mean, & SD)

***Regression Model Technique***

Given that our predictor (hospital readmission) is a binary variable, we
will perform a logistic regression.

### Data Set 2: Indicators of Heart Disease (2022)

***Response Variable: BMI***

```{r}
#| include: FALSE

# Prepping data and variables
disease_data <- readRDS("../disease_data/disease-prepared.RDS")

disease_data <- disease_data %>% 
  filter(!is.na(bmi))

bmi_table <- summary(disease_data$bmi, na.rm = TRUE)

bmi_summary <- data.frame(
  Statistic = names(bmi_table),
  BMI = as.numeric(bmi_table)
)
```

<insert text here whenever ready>

```{r}
#| echo: FALSE
#| label: fig-bmi-sum
#| tbl-cap: "Summary Statistics for BMI"

# Summary Statistics of BMI
bmi_summary %>% 
  kable(
    col.names = c("Statistic", "BMI")
  )
```

```{r}
#| echo: FALSE
#| label: fig-bmi-hist

disease_data %>% 
  ggplot() +
  aes(x = bmi) +
  geom_histogram(bins = 25, na.rm = TRUE) +
  geom_vline(xintercept = mean(disease_data$bmi, na.rm = TRUE),
             color = "tomato") +
  labs(
    x = "BMI",
    y = "Count"
  ) +
  theme_classic()
```

```{r}
# changing bmi to categorical variable
disease2 <- disease_data %>% 
  mutate(bmi = case_when(
    bmi < 19 ~ "Underweight",
    bmi >= 19 & bmi < 25 ~ "Normal Weight",
    bmi >= 25 & bmi < 30 ~ "Overweight",
    bmi >= 30 ~ "Obese"
  ))

table(disease2$bmi)

# assign counts to bmi
bmi_underweight <- disease2 %>% 
  filter(bmi == "Underweight") %>% 
  count()
bmi_under_prop <- round(bmi_underweight/nrow(disease2)*100, 2)

bmi_normal <- disease2 %>% 
  filter(bmi == "Normal Weight") %>% 
  count()
bmi_norm_prop <- round(bmi_normal/nrow(disease2)*100, 2)

bmi_overweight <- disease2 %>% 
  filter(bmi == "Overweight") %>% 
  count()
bmi_over_prop <- round(bmi_overweight/nrow(disease2)*100, 2)

bmi_obese <- disease2 %>% 
  filter(bmi == "Obese") %>% 
  count()
bmi_obese_prop <- round(bmi_obese/nrow(disease2)*100, 2)
```

```{r}
#| echo: FALSE
#| label: tbl-bmi-freq
#| tbl-cap: "BMI Category Frequency"

disease2 %>% 
  count(bmi) %>% 
  mutate(percentage = (n/nrow(disease2))*100) %>% 
  mutate(percentage = round(percentage, 2)) %>% 
  kable(col.names = c("BMI Category", "Count", "Percentage (%"))
```

- response
variable: - description - visualizations - summary statistics - 1-3
important predictors: - description - visualizations - summary
statistics - other available predictors (list them in a paragraph &
provide range of values, mean, & SD) - regression model technique -
indicate which model coefficients will be important to research question
