---
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| include: false

library(tidyverse)
library(tidyr)
library(kableExtra)
```

## 2. Exploratory Data Analysis & Analysis Approach

### Data Set 1: Diabetic Patient Hospital Readmission in the US

***Response Variable: Hospital Readmission***

```{r}
#| include: false
# PREPPING DATA & VARIABLES
diabetes_data <- readRDS("../diabetes_data/diabetes_data-prepared.RDS")

# give meaningful label to the readmitted binary variable
diabetes_data$readmitted_binary <- factor(
  diabetes_data$readmitted_binary,
  levels = c(0, 1),
  labels = c("Late Readmission (>30 Days)",
             "Early Readmission (<30 Days)")
)

# assign counts to variables
less_than_thirty<- diabetes_data |>
  filter(readmitted == "<30") |>
  count()
less_than_thirty_prop<- round(less_than_thirty/nrow(diabetes_data)*100, 2)

greater_than_thirty <- diabetes_data |>
  filter(readmitted == ">30") |>
  count()
greater_than_thirty_prop <- round(greater_than_thirty/nrow(diabetes_data)*100, 2)

no_readmission <- diabetes_data |>
  filter(readmitted == "NO") |>
  count()
no_readmission_prop <- round(no_readmission/nrow(diabetes_data)*100, 2)

late_readmission <- greater_than_thirty + no_readmission
late_readmission_prop <- round(late_readmission/nrow(diabetes_data)*100,2)

```

We will take the hospital inpatient readmission indicator (`readmitted`)
as our primary predictory variable. In our cleaned dataset, this binary
variable indicates whether a diabetic patient was a) readmitted within
30 days of their first admission, i.e. early readmission or b)
readmitted \>30 days after their first admission or not at all, i.e.
late readmission.

The vast majority of diabetic patients (n = `r late_readmission`;
 `r late_readmission_prop`% of all patients) were readmitted only after a
month (n = `r greater_than_thirty`; `r greater_than_thirty_prop`% of all
patients) or were never readmitted (n = `r no_readmission`;
 `r no_readmission_prop`% of all patients). The remaining
`r less_than_thirty` (`r less_than_thirty_prop`%) patients were readmitted
within 30 days of their first admission (@fig-readmission-freq;
@tbl-readmission-freq).

::: {.layout-ncol=2}
```{r}
#| include: false

# readmitted - NONBINARY
ggplot(data = diabetes_data, aes(x = factor(readmitted), fill = readmitted)) +
  geom_bar(color = "black") + 
  labs(x = "Hospital Readmission Status",
       y = "Count") +
  scale_fill_manual(values = c("<30" = "tomato", ">30" = "skyblue", "NO" = "steelblue")) +
  theme_classic()

```

```{r}
#| echo: false
#| label: fig-readmission-freq
# readmitted - BINARY
ggplot(data = diabetes_data, aes(x = factor(readmitted_binary), fill = readmitted)) +
  geom_bar(color = "black") + 
  labs(x = "Hospital Readmission Status",
       y = "Count") +
  #scale_x_discrete(labels = c("0" = "Late Readmission (>30 Days)",
  #                            "1" = "Early Readmission (<30 Days)")) +
  scale_fill_manual(values = c("<30" = "tomato", ">30" = "skyblue", "NO" = "steelblue")) +
  theme_classic()
```

```{r}
#| echo: false
#| label: tbl-readmission-freq
#| tbl-cap: "Hospital Readmission Frequency for Diabetic Patients in the US"

# table of readmission counts - NONBINARY
diabetes_data |> 
  count(readmitted) |>
  mutate(percentage = (n/nrow(diabetes_data))*100) |>
  mutate(percentage = round(percentage, 2)) |>
  kable(col.names = c("Readmission Status", "Count", "Percentage (%)"))

# table of readmission counts - BINARY
diabetes_data |> 
  count(readmitted_binary) |>
  mutate(percentage = (n/nrow(diabetes_data))*100) |>
  mutate(percentage = round(percentage, 2)) |>
  kable(col.names = c("Readmission Status", "Count", "Percentage (%)"))
```
:::

***Key Predictors: HbA1c test, number of diagnoses, serum glucose, drug
tests, etc.***

[Hemoglobin A1c & Serum Glucose Test Administration]{.underline}

We initially planned on investigating two categorical predictor
variables: whether a hemoglobin A1c test was administered and whether a
serum glucose test was administered. However, based on visual inspection
of the proportion of HbA1c and serum glucose test administrations for
late vs early readmissions, there does not appear to be a large
difference ( HbA1c: @fig-readmission-hba1c; Serum glucose:
@fig-readmission-serum-glu).

```{r}
#| include: false
# give meaningful label to the HbA1c binary variable
diabetes_data$a1cresult_binary <- factor(
  diabetes_data$a1cresult_binary,
  levels = c(0, 1),
  labels = c("No Test Administered",
             "Test Administered")
)

diabetes_data$a1cresult_binary
```

```{r}
#| echo: false
#| label: fig-hba1c-test

# Hemoglobin A1c Test Results
table(diabetes_data$a1cresult)

ggplot(data = diabetes_data, aes(x = factor(a1cresult_binary), fill = a1cresult)) +
  geom_bar(color = "black") + 
  labs(x = "Hemoglobin A1c Test Result",
       y = "Count") +
  scale_fill_manual(values = c(">7" = "orange", ">8" = "red",  "Norm" = "springgreen2", "None" = "gray")) +
  theme_classic()
```

```{r}
#| include: false
# give meaningful label to the serum glucose binary variable
diabetes_data$max_glu_serum_binary <- factor(
  diabetes_data$max_glu_serum_binary,
  levels = c(0, 1),
  labels = c("No Test Administered",
             "Test Administered")
)
```

```{r}
#| echo: false
#| label: fig-serum-gluc

table(diabetes_data$max_glu_serum)

# Serum Glucose Test Results

ggplot(data = diabetes_data, aes(x = factor(max_glu_serum_binary), fill = max_glu_serum)) +
  geom_bar(color = "black") + 
  labs(x = "Serum Glucose Test Result",
       y = "Count") +
  scale_fill_manual(values = c(">200" = "orange", ">300" = "red",  "Norm" = "springgreen2", "None" = "gray")) +
  theme_classic()
```

```{r}
#| echo: false
#| label: fig-readmission-hba1c

# Readmission & HbA1c Test Administration
ggplot(data = diabetes_data, aes(x = factor(readmitted_binary), fill = factor(a1cresult_binary))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Hospital Readmission Status",
       y = "Count") +
  scale_fill_manual(values = c("No Test Administered" = "tomato", "Test Administered" = "steelblue")) +
  theme_classic()
```

```{r}
#| echo: false
#| label: fig-readmission-serum-glu

# Readmission & Serum Glucose Test Administration
ggplot(data = diabetes_data, aes(x = factor(readmitted_binary), fill = factor(max_glu_serum_binary))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Hospital Readmission Status",
       y = "Count") +
  scale_fill_manual(values = c("No Test Administered" = "tomato", "Test Administered" = "steelblue")) +
  theme_classic()
```

[Number of Lab Tests & Medications Administered]{.underline}

We also planned to investigate two numerical predictor variables: 1) the
number of lab tests administered on the patient in the hospital, and 2)
the number of unique medications they were administered. However,
initial visualization of these variables as density plots and box plots
reveals that there is very little difference in the distribution and
medians of these variables for late vs early readmission groups. For
instance, the median number of lab procedures administered for patients
who were eventually readmitted late vs early was 1 for both. Likewise,
the mean was 1.4 for both (@tbl-stats-num-procedures). The median number
of medications administered to a patient was 14 for those who were
readmitted to the hospital over 30 days later and 15 for those who were
readmitted within a month. The means for these groups were 15.6 ± 8.3
and 16.6 ± 8.3, respectively — a difference that is well within the
standard deviation of both groups (@tbl-stats-num-medications).

```{r}
#| echo: false
#| label: tbl-stats-num-procedures
#| tbl-cap: "Number of Lab Procedures Administered"

# Summary Statistics - Num Lab Procedures
num_procedures_stats <- diabetes_data |>
  group_by(readmitted_binary) |>
  summarise(
    count = n(),
    mean = mean(num_procedures, na.rm = TRUE),
    median = median(num_procedures, na.rm = TRUE),
    min = min(num_procedures, na.rm = TRUE),
    max = max(num_procedures, na.rm = TRUE),
    sd = sd(num_procedures, na.rm = TRUE)
  )

# make table
num_procedures_stats |> kable(col.names = c("Readmission Status", "Count", "Mean", "Median", "Min", "Max", "SD"))

```

```{r}
#| echo: false
#| label: tbl-stats-num-medications
#| tbl-cap: "Number of Medications Administered"

# Summary Statistics - Num Medications 
num_meds_stats <- diabetes_data |>
  group_by(readmitted_binary) |>
  summarise(
    count = n(),
    mean = mean(num_medications, na.rm = TRUE),
    median = median(num_medications, na.rm = TRUE),
    min = min(num_medications, na.rm = TRUE),
    max = max(num_medications, na.rm = TRUE),
    sd = sd(num_medications, na.rm = TRUE)
  )

# make table
num_meds_stats |> kable(col.names = c("Readmission Status", "Count", "Mean", "Median", "Min", "Max", "SD"))
```

```{r, include=FALSE}
#| eval: false
#| include: false

# Histogram for number of lab tests performed
ggplot(data = diabetes_data, aes(x = num_lab_procedures, fill = readmitted_binary)) +
  geom_histogram(bins = 20, color = "black") + 
  labs(x = "Lab Procedures Administered",
       y = "Count") +
  theme_classic()
```

```{r}
#| echo: false
#| label: fig-lab-tests-freq

# Density plot
ggplot(data = diabetes_data, aes(x = num_lab_procedures, fill = readmitted_binary)) +
  geom_density(alpha = 0.5, color = "black") + 
  labs(x = "Lab Procedures Administered",
       y = "Count",
       fill = "Readmission Status") +
  theme_classic() +
  theme(legend.position = "bottom")
```

```{r}
#| echo: false
#| label: fig-lab-tests-boxplot

# Side-by-Side Boxplots
ggplot(data = diabetes_data, aes(y = num_lab_procedures, x = readmitted_binary, 
                            fill = readmitted_binary)) +
  geom_boxplot() + 
  labs(x = "Hospital Readmission Status", y = "Number of Lab Procedures Administered") + 
  #scale_fill_manual(values = c("Adelie" = "white",
  #                             "Chinstrap" = "black",
   #                            "Gentoo" = "grey")) +
  theme_classic() +
  theme(legend.position = "none")
```

```{r}
#| eval: false
#| include: false

# Histogram for medications administered
ggplot(data = diabetes_data, aes(x = num_medications, fill = readmitted_binary)) +
  geom_histogram(bins = 20, color = "black") + 
  labs(x = "Medications Administered",
       y = "Count", 
       fill = "Readmission Status") +
  theme_classic()

```

```{r}
#| echo: false
#| label: fig-medications-freq

# Density plot
ggplot(data = diabetes_data, aes(x = num_medications, fill = readmitted_binary)) +
  geom_density(alpha = 0.5, color = "black") + 
  labs(x = "Lab Procedures Administered",
       y = "Count") +
  theme_classic() +
  theme(legend.position = "bottom")
```

```{r}
#| echo: false
#| label: fig-medications-boxplot

# Side-by-Side Boxplots
ggplot(data = diabetes_data, aes(y = num_medications, x = readmitted_binary, 
                            fill = readmitted_binary)) +
  geom_boxplot() + 
  labs(x = "Hospital Readmission Status", y = "Number of Medications Administered") + 
  #scale_fill_manual(values = c("Adelie" = "white",
  #                             "Chinstrap" = "black",
   #                            "Gentoo" = "grey")) +
  theme_classic() +
  theme(legend.position = "none")
```

We then inspected the binary predictor variable `change`, indicating
whether there was a change in the patient's diabetic medications (i.e.
change in dosage or generic name). Once again, the difference
proportions for the late vs early readmission groups appears quite
small, though a slightly greater proportion of late readmission patients
(55.47%) experienced no change in medications compared to early
admission patients (52.66%) (@fig-changed-medication). 

Interestingly, it appears that a larger proportion of patients who were administered an HbA1c test also received a change in medication (55.04%), compared to patients who were not given an HbA1c test (47.34%) (@fig-hba1c-med-change).

Further statistical analysis would be required to determine significance.

```{r}
#| echo: false
#| label: fig-changed-medication
ggplot(data = diabetes_data, aes(x = factor(readmitted_binary), fill = change)) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Hospital Readmission Status",
       y = "Proportion",
       fill = "Change in Medication") +
  scale_fill_manual(
    values = c("No" = "tomato", "Ch" = "steelblue"),
    labels = c("No" = "No", "Ch" = "Yes")) +

  theme_classic()



```

```{r}
#| include: false
late_readmission <- diabetes_data |> filter(readmitted_binary == "Late Readmission (>30 Days)")

late_readmission |> count(change) |>
  mutate(percentage = (n/nrow(late_readmission))*100) |>
  mutate(percentage = round(percentage, 2)) |>
  kable(col.names = c("Change", "N", "Percentage (%)"))
# change: 44.53%
# no change: 55.47%

early_readmission <- diabetes_data |> filter(readmitted_binary == "Early Readmission (<30 Days)")

early_readmission |> count(change) |>
  mutate(percentage = (n/nrow(early_readmission))*100) |>
  mutate(percentage = round(percentage, 2)) |>
  kable(col.names = c("Change", "N", "Percentage (%)"))

# change: 47.34%
# no change: 52.66%
```



```{r}
#| echo: false
#| label: fig-hba1c-med-change

ggplot(data = diabetes_data, aes(x = factor(a1cresult_binary), fill = change)) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "HbA1c Result",
       y = "Proportion",
       fill = "Change in Medication") +
  scale_fill_manual(
    values = c("No" = "tomato", "Ch" = "steelblue"),
    labels = c("No" = "No", "Ch" = "Yes")) +

  theme_classic()

```

```{r}
#| include: false
yes_hba1c <- diabetes_data |> filter(a1cresult_binary == "Test Administered")

yes_hba1c |> count(change) |>
  mutate(percentage = (n/nrow(yes_hba1c))*100) |>
  mutate(percentage = round(percentage, 2)) |>
  kable(col.names = c("Change", "N", "Percentage (%)"))
# change: 55.04%
# no change: 44.96%

no_hba1c <- diabetes_data |> filter(a1cresult_binary == "No Test Administered")

no_hba1c |> count(change) |>
  mutate(percentage = (n/nrow(no_hba1c))*100) |>
  mutate(percentage = round(percentage, 2)) |>
  kable(col.names = c("Change", "N", "Percentage (%)"))

# change: 42.5%
# no change: 57.5%
```


Overall, investigating the influence of all five of these predictors
(HbA1c testing, serum glucose testing, number of lab tests administered,
number of medications administered, and change in medication) may not
yield fruitful results in future analyses given the above
visualizations. However, we cannot make any clear conclusions at this stage.



***Other Predictors***

We originally also considered looking at the number of non-lab
procedures performed during the patient's hospital visit and the number
of outpatient visits in the year prior to the recorded hospital
encounter. However, after visualizing these variables as histograms
(@fig-outpatient-freq; @fig-nolab-tests-freq), these predictor variables
are highly abnormally distributed. As such, we will likely omit these
variables from future analyses.

```{r}
#| echo: false
#| label: fig-nolab-tests-freq

# Histogram for number of lab tests performed
ggplot(data = diabetes_data, aes(x = num_procedures)) +
  geom_histogram(bins = 10, color = "black", fill = "white") + 
  labs(x = "Non-Lab Procedures Administered",
       y = "Count") +
  theme_classic()
```

```{r}
#| echo: false
#| label: fig-outpatient-freq

# Histogram for outpatient visits in year prior to encounter
ggplot(data = diabetes_data, aes(x = num_outpatient_visits)) +
  geom_histogram(bins = 20, color = "black", fill = "white") + 
  labs(x = "Outpatient Visits in Past Year",
       y = "Count") +
  theme_classic()
```

We then performed initial visualizations for all other variables, which
included race, gender, and the 23 different drugs that hospitals may
have administered. All variables yielded similarly unpromising plots to
those shown above for serum glucose and HbA1c.

The duration of the hospital stay was the only other numerical predictor
of potential interest. The summary statistics are given in
@tbl-time-in-hospital. Once again, the mean and medians are very similar
for both late and early readmission groups.

```{r}
#| echo: false
#| label: tbl-time-in-hospital
#| tbl-cap: Days Spent in the Hospital

# Time spent in hospital
diabetes_data |>
  group_by(readmitted_binary) |>
  summarise(
    count = n(),
    mean = mean(time_in_hospital, na.rm = TRUE),
    median = median(time_in_hospital, na.rm = TRUE),
    min = min(time_in_hospital, na.rm = TRUE),
    max = max(time_in_hospital, na.rm = TRUE),
    sd = sd(time_in_hospital, na.rm = TRUE)
  ) |>
  kable(col.names = c("Readmission Status", "Count", "Mean", "Median", "Min", "Max", "SD"))


```

```{r}
#| include: false
#| label: fig-time-in-hospital

# Histogram for number of lab tests performed
ggplot(data = diabetes_data, aes(x = time_in_hospital)) +
  geom_histogram(bins = 20, color = "black", fill = "white") + 
  labs(x = "Days Between Admission and Discharge ",
       y = "Count") +
  theme_classic()
```

***Regression Model Technique***

If we were to move forward with inferential analysis, we would perform a
logistic regression given that our predictor (hospital readmission) is a
binary variable, . The coefficients of interest would be 1)
administration of HbA1c test, 2) administration of serum glucose test,
and 3) change in medications.


### Data Set 2: Indicators of Heart Disease (2022)

***Response Variable: BMI***

In addition to our primary binary response variable (`readmitted`), we will take the variable `bmi` as our another primary response variable. In our cleaned dataset, this continuous variable provides a quantitative measure of body mass that we can use to explore patterns, variability, and potential relationships with other health-related factors.

BMI ranged from as low as 15, indicating underweight, to nearly 60, reflecting obesity. On average, individuals in the United States have a BMI around 28, and the median is slightly lower at 27.3 (@tbl-bmi-stat). This difference between the mean and median reflects a right-skewed distribution. Most individuals have BMI values near the mean, but a smaller number have much higher BMI values, reflecting obesity (@fig-bmi-hist).

```{r}
#| include: false

# Prepping data and variables
disease_data <- readRDS("../disease_data/disease-prepared.RDS")

disease_data <- disease_data %>% 
  filter(!is.na(bmi))
```

```{r}
#| echo: FALSE
#| label: tbl-bmi-stat
#| tbl-cap: "Summary Statistics for BMI"

bmi_stats <- disease_data %>% 
  summarise(
    mean = mean(bmi, na.rm = TRUE),
    median = median(bmi, na.rm = TRUE),
    min = min(bmi, na.rm = TRUE),
    max = max(bmi, na.rm = TRUE),
    sd = sd(bmi, na.rm = TRUE)
  )

# make table
bmi_stats %>% 
  kable(
    col.names = c("Mean", "Median", "Min", "Max", "SD")
  )
```

```{r}
#| echo: FALSE
#| label: fig-bmi-hist

disease_data %>% 
  ggplot() +
  aes(x = bmi) +
  geom_histogram(bins = 25, na.rm = TRUE, fill = "springgreen2", color = "black") +
  geom_vline(xintercept = mean(disease_data$bmi, na.rm = TRUE),
             color = "red") +
  labs(
    x = "BMI",
    y = "Count"
  ) +
  theme_classic()
```

Only 2.27% of individuals in the United States have a BMI classified as underweight. About 29.75% have a BMI in the normal range, 35.90% are overweight, and 32.08% are classified as obese (@tbl-bmi-freq; @fig-bmi-bar).

```{r}
#| include: FALSE

# changing bmi to categorical variable
disease2 <- disease_data %>% 
  mutate(bmi = case_when(
    bmi < 19 ~ "Underweight",
    bmi >= 19 & bmi < 25 ~ "Normal Weight",
    bmi >= 25 & bmi < 30 ~ "Overweight",
    bmi >= 30 ~ "Obese"
  ))

table(disease2$bmi)

# ordering bmi categories
disease2 <- disease2 %>% 
  mutate(
    bmi = factor(
      bmi, levels = c("Underweight", "Normal Weight", "Overweight", "Obese")
    )
  )

# assign counts to bmi
bmi_underweight <- disease2 %>% 
  filter(bmi == "Underweight") %>% 
  count()
bmi_under_prop <- round(bmi_underweight/nrow(disease2)*100, 2)

bmi_normal <- disease2 %>% 
  filter(bmi == "Normal Weight") %>% 
  count()
bmi_norm_prop <- round(bmi_normal/nrow(disease2)*100, 2)

bmi_overweight <- disease2 %>% 
  filter(bmi == "Overweight") %>% 
  count()
bmi_over_prop <- round(bmi_overweight/nrow(disease2)*100, 2)

bmi_obese <- disease2 %>% 
  filter(bmi == "Obese") %>% 
  count()
bmi_obese_prop <- round(bmi_obese/nrow(disease2)*100, 2)
```

```{r}
#| echo: FALSE
#| label: tbl-bmi-freq
#| tbl-cap: "BMI Category Frequency"

disease2 %>% 
  count(bmi) %>% 
  mutate(percentage = (n/nrow(disease2))*100) %>% 
  mutate(percentage = round(percentage, 2)) %>% 
  kable(col.names = c("BMI Category", "Count", "Percentage (%)"))
```
```{r}
#| echo: FALSE
#| label: fig-bmi-bar

disease2 %>% 
  ggplot() +
  aes(x = bmi, fill = bmi) +
  geom_bar() +
  labs(
    x = "BMI Category",
    y = "Count",
    fill = "Category"
  ) +
  scale_fill_manual(values = c(
    "Underweight" = "orange",
    "Normal Weight" = "steelblue",
    "Overweight" = "skyblue",
    "Obese" = "tomato"
  )) +
  theme_classic()
  
```

***Key Predictors: Smoking, Physical Activity, Sleep Time, and Age***

[Smoking, Age, and Physical Activity]{.underline}

**Insert third text whenever ready.**

We also plan to investigate three categorical predictors: smoking status, physical activity, and age. Over 55% of the individuals surveyed were 55 years or older (@tbl-age-freq; @fig-age). In terms of phsyical activity, more than 77% of individuals were active, while 22.41% were inactive (@tbl-physical-freq; @fig-physical).

```{r}
#| echo: FALSE

# giving meaningful label to smoking

disease2$smoking <- disease_data$smoking

disease2$smoking <- factor(
  disease2$smoking,
  levels = c(0, 1),
  labels = c("No", "Yes")
)
```

```{r}
#| echo: FALSE
#| label: tbl-smoke-freq
#| tbl-cap: "Smoking Status Frequency"

disease2 %>% 
  count(smoking) %>% 
  mutate(percentage = (n/nrow(disease2))*100) %>% 
  mutate(percentage = round(percentage, 2)) %>% 
  kable(col.names =
          c("Smoking Status", "Count", "Percentage(%)"))
```

```{r}
#| echo: FALSE
#| label: fig-smoke

disease2 %>% 
  ggplot() +
  aes(x = smoking, fill = smoking) +
  geom_bar() +
  labs(
    x = "Smoking Status",
    y = "Count",
    fill = "Status"
  ) +
  scale_fill_manual(values = c(
    "Yes" = "tomato",
    "No" = "skyblue"
  )) +
  theme_classic()
```

```{r}
#| echo: FALSE

# giving meaningful label to physical_activity

disease2$physical_activity <- disease_data$physical_activity

disease2$physical_activity <- factor(
  disease2$physical_activity,
  levels = c(0, 1),
  labels = c("Not Active", "Active")
)
```

```{r}
#| echo: FALSE
#| label: tbl-physical-freq
#| tbl-cap: "Physical Activity Frequency"

disease2 %>% 
  count(physical_activity) %>% 
  mutate(percentage = (n/nrow(disease2))*100) %>% 
  mutate(percentage = round(percentage, 2)) %>% 
  kable(col.names =
          c("Physical Activity", "Count", "Percentage(%)"))
```

```{r}
#| echo: FALSE
#| label: fig-physical

disease2 %>% 
  ggplot() +
  aes(x = physical_activity, fill = physical_activity) +
  geom_bar() +
  labs(
    x = "Physical Activity",
    y = "Count",
    fill = "Status"
  ) +
  scale_fill_manual(values = c(
    "Active" = "tomato",
    "Not Active" = "skyblue"
  )) +
  theme_classic()
```

```{r}
# echo: FALSE

# reassigning age-ranges for efficiency
# 18-34, 35-49, 50-64, 65 and older
disease2$age_category <- disease_data$age_category

disease2 <- disease2 %>% 
  mutate(age_category = case_when(
    age_category %in% c("18-24", "25-29", "30-34") ~ "18-34",
    age_category %in% c("35-39", "40-44", "45-49", "50-54") ~ "35-54",
    age_category %in% c("55-59", "60-64", "65-69", "70-74", "75-79", "80 or older") ~ "55 and older"
  ))
```

```{r}
#| echo: FALSE
#| label: tbl-age-freq
#| tbl-cap: "Age Range Frequency"

disease2 %>% 
  count(age_category) %>% 
  mutate(percentage = (n/nrow(disease2))*100) %>% 
  mutate(percentage = round(percentage, 2)) %>% 
  kable(col.names =
          c("Age Range", "Count", "Percentage(%)"))
```

```{r}
#| echo: FALSE
#| label: fig-ages

disease2 %>% 
  ggplot() +
  aes(x = age_category, fill = age_category) +
  geom_bar() +
  labs(
    x = "Age Range",
    y = "Count",
    fill = "Age Range"
  ) +
  scale_fill_manual(values = c(
    "18-34" = "tomato",
    "35-54" = "skyblue",
    "55 and older" = "orange"
  )) +
  theme_classic()
```

- response
variable: - description - visualizations - summary statistics - 1-3
important predictors: - description - visualizations - summary
statistics - other available predictors (list them in a paragraph &
provide range of values, mean, & SD) - regression model technique -
indicate which model coefficients will be important to research question
