---
editor: 
  markdown: 
    wrap: 72
---

## Methodology

<!-- If you use a single dataset, use this structure -->

### Data

```{r}
#| include: false
# loading libraries

library(tidyverse)
library(tidyr)
library(kableExtra)
library(bestglm)
library(car)

```

```{r}
#| include: false
# loading data
heartdisease <- readRDS("../disease_data/heart-prepared.RDS")

# relevel data to set baselines for some predictors
heartdisease$race <- relevel(factor(heartdisease$race), ref = "White")

heartdisease$age_category <- relevel(factor(heartdisease$age_category), ref = "1")

heartdisease$gen_health <- relevel(factor(heartdisease$gen_health), ref = "Poor")

heartdisease$age_category <- relevel(factor(heartdisease$age_category), ref = "1")

```

The current dataset on crucial indicators for heart disease from
patients in the United States (2022) is sourced from the following
website hosted by Kaggle:
<https://www.kaggle.com/datasets/kamilpytlak/personal-key-indicators-of-heart-disease>.
The dataset originally comes from the Centers for Disease Control and
Prevention (CDC) and is a part of the Behavioral Risk Factor
Surveillance System (BRFSS). It was originally collected by the CDC and
BRFSS to gather data on the health status of U.S. residents across all
50 states, the District of Columbia, and 3 U.S. territories. To download
the dataset, scroll to the end of the "About Dataset" section and click
the download icon (an arrow pointing down). The dataset was downloaded
on February 22, 2026.

The raw dataset contains 319795 observations, each representing an
individual surveyed through the BRFSS in 2022. Each individual answered
a series of questions regarding their health. Eighteen variables were
recorded for each individual, including identity characteristics (e.g.,
sex and race) and indicators of overall cardiovascular health, such as
BMI, smoking status, mental health, and diabetes status.

The original purpose of the dataset was to monitor the population health
and identify crucial risk factors for diseases such as heart disease to
inform public health policy and prevention efforts. This research builds
on that goal by providing further insight into how these risk factors
contribute to the prediction of heart disease and overall cardiovascular
health outcomes.

The dataset was cleaned by addressing implausible values in numeric
variables, specifically `bmi` and `sleep_time`. In the `bmi` variable,
extremely low values were observed (minimum = 12.02) that are considered
physiologically implausible; these values were treated as NA. Similarly,
the `sleep_time` variable included extreme values (minimum = 1 hour and
maximum = 24 hours). While achieving such durations is technically
possible, they are highly unusual and could disproportionately influence
subsequent analyses, so they were also recorded as NA to improve data
quality and reliability. In addition, the `diabetic` variable was
altered to have only two levels ("Yes" and "No") instead of its original
four levels ("Yes", "No", "Yes (during pregnancy)", and "No, borderline
diabetes"). This simplifies future analyses, where we will only be
concerned whether diabetes was present, regardless of when or in what
context.



### Exploratory Data Analyses

#### BMI and Predictors

The distribution of BMI (@fig-bmi-hist) shows a strong right-skew. Most
observations are clustered between 25 and 30, but a few outliers extend
to 60. This non-normality suggests that a standard linear model might be
biased by extreme values.

```{r}
#| echo: false
#| label: fig-bmi-hist

heartdisease %>% 
  ggplot() +
  aes(x = bmi) +
  geom_histogram(na.rm = TRUE, fill = "skyblue", color = "black") +
  geom_vline(xintercept = mean(heartdisease$bmi, na.rm = TRUE),
             color = "red") +
  labs(
    x = "BMI",
    y = "Count"
  ) +
  theme_classic()

# bmi is slightly right-skewed

# log(bmi) is less skewed
```

```{r}
#| include: false

summary(heartdisease$bmi)
# min = 15
# median = 27.32
# mean = 28.28
# max = 59.97
```

The boxplot of BMI across age groups (@fig-bmi-age) reveals that the
median BMI is highest in the middle-age group (35-54) and slightly lower
in the youngest (18-34) and oldest (55+) cohorts. While the
interquartile ranges are similar across groups, all three groups exhibit
a massive density of high-value outliers, with BMI values extending up
to 60. This suggests that extreme BMI values are prevalent regardless of
age.

```{r}
#| echo: false
#| label: fig-bmi-age

heartdisease %>% 
  filter(!is.na(age_category), !is.na(bmi)) %>%
  ggplot() +
  aes(x = age_category, y = bmi, fill = age_category) +
  geom_boxplot() +
  labs(
    x = "Age Groups",
    y = "BMI",
    fill = "Age Groups"
  ) +
  theme_classic() +
  scale_fill_manual(values = c(
    "1" = "orange",
    "2" = "tomato",
    "3" = "skyblue"),
    labels = c(
      "1" = "18-34",
      "2" = "35-54",
      "3" = "55 and older")
    ) +
  scale_x_discrete(labels = c(
  "1" = "18-34",
  "2" = "35-54",
  "3" = "55 and older"
))
```

Comparing BMI across racial groups (@fig-bmi-race) shows noticeable
variation in central tendency; Asian individuals display the lowest
median BMI and a narrower overall distribution, while Black and Native
American individuals show higher median values. Similar to the age
categories, every racial group contains a significant number of extreme
outliers, indicating a consistent right-skew in the BMI data across all
individuals.

```{r}
#| echo: false
#| label: fig-bmi-race

heartdisease %>% 
  filter(!is.na(race), !is.na(bmi)) %>%
  ggplot() +
  aes(x = race, y = bmi, fill = race) +
  geom_boxplot() +
  labs(
    x = "Race",
    y = "BMI",
    fill = "Race"
  ) +
  theme_classic() +
  scale_fill_manual(values = c(
    "White" = "antiquewhite",
    "American Indian/Alaskan Native" = "tomato",
    "Asian" = "skyblue",
    "Black" = "orange",
    "Hispanic" = "lightsalmon",
    "Other" = "plum"),
    labels = c(
      "American Indian/Alaskan Native" = "Native American"
      )
    ) +
  scale_x_discrete(labels = c(
  "American Indian/Alaskan Native" = "Native American"))
```

```{r}
#| include: false

# correlations for numeric predictors

cor(heartdisease$sleep_time, heartdisease$bmi, use = "complete.obs") # -0.05622031

cor(heartdisease$physical_health, heartdisease$bmi, use = "complete.obs") # 0.1069033

cor(heartdisease$mental_health, heartdisease$bmi, use = "complete.obs") #0.06191527
```

```{r}
#| include: false

# interactions?

aggregate(bmi ~ sleep_time, data = heartdisease, FUN = mean)

# none, consistent

```

### Statistical Methods

#### Multiple Linear Regression: Factors that Determine BMI


**Selecting a Multiple Linear Regression Model**

Given that our cleaned data set only contains 17 potential predictors,
we opted to employ all-subset, forward stepwise, and backward stepwise
model selection processes, as all three methods are not too
computationally expensive in our case. Before performing model
selection, the dataset was split into a training set (80%) and a testing
set (20%) to ensure the generalizability of the final model. A random
seed (1234) was set to ensure reproducibility of the split. The training
set (n = 255836) was used for model development and variable selection,
while the testing set was reserved for final model validation to assess
its predictive performance on unseen data.

To satisfy the linear regression assumption of normality (particularly
given the skewness observed in the raw BMI distribution earlier), a
natural log transformation of BMI (`log(bmi)`) was applied during model
selection, which stabilizes variance and ensures more reliable
coefficient estimates.

All three model selection methods identified the same predictors that
minimized BIC, indicating that these variables collectively explained
variability in `log(bmi)` effectively while appropriately penalizing
model complexity and reducing the risk of overfitting. These variables
were the following: `smoking`, `alcohol_drinking`, `stroke`,
`physical_health`, `diff_walking`, `sex`, `age_category` (particularly
ages 35 to 54 years), `race` (particularly Asian and Black), `diabetic`,
`gen_health`, `sleep_time`, `asthma`, and `skin_cancer`.

```{r}
#| include: false

set.seed(1234)

# 80% of data
n <-nrow(heartdisease)
train_indices <- sample(1:n, size = round(0.80 * n))

# add indices to data
heartdisease <- heartdisease %>% mutate(
  row_id = row_number(),
  set = ifelse(row_id %in% train_indices, "train", "test")
)

# filter train data
heart_train <- heartdisease %>% 
  filter(set == "train") %>% 
  select(-row_id, -set)

# filter test data
heart_test <- heartdisease %>% 
  filter(set == "test") %>% 
  select(-row_id, -set)
```

```{=html}
<!-- 
Method 1: MLR All Subset Selection
-->
```

```{r}
#| include: false
# all-subset selection

best_mlr_all_sub <- regsubsets(log(bmi) ~ .,
                         nvmax = 17,
                         data = heart_train)

mlr_model_sum <- summary(best_mlr_all_sub)
mlr_best_idx <- which.min(mlr_model_sum$bic) # find the index in that corresponds to the model with
                                          # the lowest BIC
selected_mlr <- mlr_model_sum$which[mlr_best_idx, ] # select the row best_idx in the logical matrix --> selected_model is a logical vector
selected_mlr |> kable() 

selected_mlr_vars <- names(selected_mlr)[selected_mlr] # keeping only the TRUE predictors

```

::::: {layout="[0.25, 0.75]"}
::: {#plot1}
```{r}
#| include: false
#| label: tbl-all-sub-set-mlr
#| tbl-cap: "Exhaustive MLR Model Selection: Selected Predictors Based on BIC"

selected_mlr_vars |> kable()
```
:::

::: {#plot2}
```{r}
#| include: false
#| label: fig-all-subset-mlr
#| fig-cap: "Exhaustive Model Selection, B) Forward Model Selection"
plot(best_mlr_all_sub, scale = "bic")

```
:::
:::::

```{=html}
<!-- 
Method 2: Forward Selection
-->
```

```{r}
#| include: false
# forward selection

best_mlr_fwd <- regsubsets(log(bmi) ~ .,
                         method = "forward",
                         nvmax = 17,
                         data = heart_train)

mlr_model_sum <- summary(best_mlr_fwd)
mlr_best_idx <- which.min(mlr_model_sum$bic) # find the index in that corresponds to the model with
                                          # the lowest BIC
selected_mlr <- mlr_model_sum$which[mlr_best_idx, ] # select the row best_idx in the logical matrix --> selected_model is a logical vector
selected_mlr |> kable() 

selected_mlr_vars <- names(selected_mlr)[selected_mlr] # keeping only the TRUE predictors
selected_mlr_vars |> kable()

```

::::: {layout="[0.25, 0.75]"}

```{r}
#| include: false
#| label: tbl-fwd-mlr
#| tbl-cap: "Forward Model Selection: Selected Predictors Based on BIC"

selected_mlr_vars |> kable()
```



```{r}
#| include: false
#| label: fig-fwd-mlr
#| fig-cap: "Forward Model Selection (BIC)"

plot(best_mlr_fwd, scale = "bic")
```

</div>
:::::

```{=html}
<!-- 
Method 3: Backward Selection
-->
```

```{r}
#| include: false
# backward selection

best_mlr_bwd <- regsubsets(log(bmi) ~ .,
                         method = "backward",
                         nvmax = 17,
                         data = heart_train)

mlr_model_sum <- summary(best_mlr_bwd)
mlr_best_idx <- which.min(mlr_model_sum$bic) # find the index in that corresponds to the model with
                                          # the lowest BIC
selected_mlr <- mlr_model_sum$which[mlr_best_idx, ] # select the row best_idx in the logical matrix --> selected_model is a logical vector
selected_mlr |> kable() 

selected_mlr_vars <- names(selected_mlr)[selected_mlr] # keeping only the TRUE predictors
selected_mlr_vars |> kable()

```

::::: {layout="[0.25, 0.75]"}
<div>

```{r}
#| include: false
#| label: tbl-bwd-mlr
#| tbl-cap: "Bakward Model Selection: Selected Predictors Based on BIC"

selected_mlr_vars |> kable()
```

</div>

<div>

```{r}
#| include: false
#| label: fig-bwd-mlr
#| fig-cap: "Backward Model Selection (BIC)"

plot(best_mlr_bwd, scale = "bic")
```

</div>
:::::

::::: {layout="[0.25, 0.75]"}
<div>

```{r}
#| echo: false
#| label: tbl-selected-vars-mlr
#| tbl-cap: "Selected Predictors Based on BIC (for exhaustive, forward, and backward selection technqiues)"

selected_mlr_vars |> kable()
```

</div>

<div>

```{r}
#| echo: false
#| fig-cap: "A) Exhaustive Model Selection"

# Exhaustive
plot(best_mlr_all_sub, scale = "bic")

```

```{r}
#| echo: false
#| fig-cap: "B) Forward Model Selection"
#| # Forward
plot(best_mlr_fwd, scale = "bic")

```

```{r}
#| echo: false
#| fig-cap: "C) Backward Model Selection"
# Backward
plot(best_mlr_bwd, scale = "bic")
```

</div>
:::::

**Validating Final Model Assumptions**

Given the results of our model selection, we moved forward with the
following multiple linear regression:

```{r}
#| include: false

chosen_mlr <- lm(log(bmi) ~ smoking + alcohol_drinking + stroke + physical_health +
                   diff_walking + sex + age_category + race + diabetic + physical_activity +
                   gen_health + sleep_time + asthma + skin_cancer, 
                 data = heart_train)

summary(chosen_mlr)
```

$$
\begin{aligned}
log(BMI_i) = \beta_0 + \beta_1Smoking_i + \beta_2Alcohol_i + \beta_3Stroke_i + \\
\beta_4PhysicalHealth_i + \beta_5DiffWalking_i + \beta_6Sex_i + \beta_7{Age_{35-54}}_i + \beta_8Asian_i + \\
\beta_9Black_i + \beta_{10}Diabetic_i + \beta_{11}HealthFair_i + \beta_{12}HealthGood_i + \\
\beta_{13}HealthExcellent_i + \beta_{14}SleepTime_i + \beta_{15}Asthma_i + \\
\beta_{16}SkinCancer_i
\end{aligned}
$$

Interestingly, the multiple linear regression yielded a p-value of
3.10e-05 (\< 0.05) for $\beta_{NativeAmerican}$, indicating that the
predictor is significantly associated with variation in log-transformed
BMI, holding all other variables constant. Given this, the predictor was
included in the final model below:

```{r}
#| include: false

final_mlr <- lm(log(bmi) ~ smoking + alcohol_drinking + stroke + physical_health +
                   diff_walking + sex + age_category + race + diabetic + physical_activity +
                   gen_health + sleep_time + asthma + skin_cancer, 
                 data = heart_train)

summary(final_mlr)
```

$$
\begin{aligned}
log(BMI_i) = \beta_0 + \beta_1Smoking_i + \beta_2Alcohol_i + \beta_3Stroke_i + \\
\beta_4PhysicalHealth_i + \beta_5DiffWalking_i + \beta_6Sex_i + \beta_7{Age_{35-54}}_i + \beta_8Asian_i + \\
\beta_9Black_i + \beta_{10}NativeAmerican_i + \beta_{11}Diabetic_i + \beta_{12}HealthFair_i + \\
\beta_{13}HealthGood_i + \beta_{14}HealthExcellent_i + \beta_{15}SleepTime_i + \beta_{16}Asthma_i + \\
\beta_{17}SkinCancer_i
\end{aligned}
$$

{{< pagebreak >}}

Linearity and homoscedasticity were evaluated using a residuals versus
fitted values plot (@fig-mlr-resfit). The assumptions are largely
satisfied, as the residuals are centered around zero with a relatively
consistent vertical spread across the range of fitted values.

Next, normality of residuals was assessed through a Q-Q plot
(@fig-mlr-qq). The majority of the standardized residuals follow the
theoretical normal line closely. While there is slight "thin-tailing" at
the extremes, the distribution of errors is sufficiently normal for
robust statistical inference, especially given the large sample size.

We also examined potential influential observations using a residuals
versus leverage plot (i.e., a Cook's distance plot) (@fig-mlr-leverage).
All values remained well below the common threshold of 1.0, indicating
that no single observation is unduly biasing the model's estimates.

::::: {layout="[0.45, 0.45]"}
::: {#plot1}
```{r}
#| echo: false
#| label: fig-mlr-resfit

plot(final_mlr, which = 1)
```
:::

:::{#plot2}
```{r}
#| echo: false
#| label: fig-mlr-qq

plot(final_mlr, which = 2)
```
:::

::::

::::: {layout="[0.2, 0.5, 0.2]"}
:::{#space}
:::
::: {#plot1}
```{r}
#| echo: false
#| label: fig-mlr-leverage
#| fig-align: center

plot(final_mlr, which = 4)
```
:::

:::{#space}
:::
::::

Lastly, we assessed multicollinearity by calculating the General
Variance Inflation Factors (GVIFs). The VIF is calculated as follows:
$VIF_j = \frac{1}{1 - R_j^2}$ and indicates to what extent a the ariance
in predictor $X_j$ can be explained by the other predictors. To avoid
collinearity, we would hope that $R_j^2$ is as close to 0 as possible,
in which case $VIF_j$ is 1. In general, a VIF score of less than or
equal to 5 is still considered acceptable for collinearity (i.e. when
$R_j^2 \le 0.2$).

All GVIF values were below 2.0, indicating that multicollinearity is not
a concern for our final model.

```{r}
#| include: false
# VIF indicates "how much variance of estimated regression coefficient is increased due 
# to collinearity with other predictors"
# VIF values of 1 indicate no correlation
# VIF >5-10 indicate high multicollinearity
vif(final_mlr)

```

{{< pagebreak >}}

#### Logistic Regression: Predicting Heart Attack

We were furthermore interested in assessing an individual's risk for
heart attack based on demographics, particularly race, behavior
(smoking, drinking alcohol, sleep, physical activity), and other health
conditions (diabetes, stroke, kidney disease, asthma). Given that
`heart_disease` is a binary outcome variable, we opted to perform a
**logistic regression**.

Initially, we investigated BMI as a potential predictor for heart
attack, as BMI is a simple and readily accessible health measurement.
Initial visualization of heart disease status and BMI did not reveal an
evident difference in BMI amongst individuals with and without heart
disease (@fig-heartdisease-bmi-hist). Categorizing BMIs into different
weight categories seemed to reveal that a slightly larger proportion of
individuals with heart disease had BMI values classified as obese
(@fig-heartdisease-bmi-bar). However, as will be shown further on in
this analysis, BMI was omitted as a predictor in our final regression
model.

```{r}
#| include: false

library(tidyverse)
library(tidyr)
library(kableExtra)
```

```{r}
#| include: false

# load data
disease_data <- readRDS("../disease_data/disease-prepared.RDS")
summary(disease_data) # only 1 numeric variable

# relevel the data to make sure that raceWhite is used as the baseline
disease_data$race <- relevel(factor(disease_data$race), ref = "White")
disease_data$gen_health <- relevel(factor(disease_data$gen_health), ref = "Poor")

heart_train$race <- relevel(factor(heart_train$race), ref = "White")
heart_train$gen_health <- relevel(factor(heart_train$gen_health), ref = "Poor")



```

::::: {layout="[0.45, 0.45]"}
::: column
```{r}
#| echo: false
#| label: fig-heartdisease-bmi-hist
## BODY MASS INDEX (BMI)

ggplot(data = disease_data, aes(x = bmi, fill = factor(heart_disease))) +
  geom_density(alpha = 0.5, color = "black") + 
  labs(x = "BMI",
       y = "Count",
       fill = "Heart Disease") +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() +
  theme(legend.position = "bottom")

```

```{r}
#| include: false

# BMI
ggplot(data = disease_data, aes(y = bmi, x = factor(heart_disease), 
                            fill = factor(heart_disease))) +
  geom_boxplot() + 
  labs(x = "Heart Disease", y = "BMI") + 
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() +
  theme(legend.position = "none")
```
:::

::: column
```{r}
#| echo: false
#| label: fig-heartdisease-bmi-bar

disease2 <- disease_data |>
  mutate(sleep_time = case_when(
    sleep_time <= 10 & sleep_time >= 7 ~ "Healthy Sleep (7-10 hrs)",
    sleep_time < 7 ~ "Sleep Deprived (<7 hrs)",
    sleep_time > 10 ~ "Extra Sleep (>10 hrs)",
    TRUE ~ NA_character_),
    
    bmi = case_when(
    bmi < 19 ~ "Underweight",
    bmi >= 19 & bmi < 25 ~ "Normal Weight",
    bmi >= 25 & bmi < 30 ~ "Overweight",
    bmi >= 30 ~ "Obese",
    TRUE ~ NA_character_),
    
   # age_category = case_when(
 #   age_category %in% young ~ "Young (18-39)",
  #  age_category %in% middle_age ~ "Middle Aged (40-59)",
  #  age_category %in% old ~ "Elderly (60-80+)",
  #  TRUE ~ NA_character_)
    
    )

# BMI
ggplot(data = disease2, aes(x = factor(heart_disease), fill = bmi)) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease",
       y = "Count",
       fill = "BMI") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  theme_classic() + 
  theme(legend.position = "bottom")
```
:::
:::::

```{r}
#| include: false

# BMI
ggplot(data = disease_data, aes(y = bmi, x = factor(heart_disease), 
                            fill = factor(heart_disease))) +
  geom_boxplot() + 
  labs(x = "Heart Disease", y = "BMI") + 
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() +
  theme(legend.position = "none")

```

::::: {layout="[0.3, 0.3, 0.3]"}
<div>

```{r}
#| include: false

# PHYSICAL ACTIVITY
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = factor(physical_activity))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Physically Active") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() + 
  theme(legend.position = "bottom")
```

</div>

<div>

```{r}
#| include: false
## BEHAVIOR

# SLEEP
ggplot(data = disease2, aes(x = factor(heart_disease), fill = sleep_time)) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease",
       y = "Count",
       fill = "Sleep") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  theme_classic() + 
  theme(legend.position = "bottom")
```

</div>
:::::

```{r}
#| include: false

# correlations for numeric predictors

cor(disease_data$sleep_time, disease_data$heart_disease, use = "complete.obs") # 0.009097552

cor(disease_data$physical_health, disease_data$heart_disease, use = "complete.obs") # 0.170721

cor(disease_data$mental_health, disease_data$heart_disease, use = "complete.obs") # 0.02859071
```

**Selecting a Logistic Regression Model**

Given that the our cleaned heart disease data set only contains 17
possible predictors, we opted to employ both a all-subset, stepwise
forward, and a backward elimination model selection process, as all
three methods are not too computationally expensive in our case. For all
models, we used the same 80%-20% split of train vs test data selected
for our multiple linear regression of BMI in the section above.

All three model selection methods yielded the same predictors that
minimized the BIC, i.e. accounted for the largest amount of variance in
the log-odds of heart disease while mitigating overfitting and
optimizing model parsimony. These variables were the following:
`smoking`,`alcohol_drinking`, `stroke`, `mental_health`, `diff_walking`,
`sex`, `age_category` (particularly ages 55 years and above), `race`
(particularly Asian, Black, and Hispanic with White as a baseline),
`diabetic`, `gen_health`, `asthma`, `kidney_disease`, and `skin_cancer`.
Notably, all three methods omitted the following predictors: `bmi`,
`physical_activity`, `physical_health`, and `sleep_time`.

```{=html}
<!-- 
Method 1: All Subset Selection
-->
```

```{r}
#| include: false
# all-subset selection

best_log_reg_all_sub <- regsubsets(heart_disease ~ .,
                         method = "exhaustive",
                         nvmax = 17,
                         data = heart_train)

model_summary <- summary(best_log_reg_all_sub)
best_idx <- which.min(model_summary$bic) # find the index in that corresponds to the model with
                                          # the lowest BIC
selected_model <- model_summary$which[best_idx, ] # select the row best_idx in the logical matrix --> selected_model is a logical vector
selected_model |> kable() 

selected_vars <- names(selected_model)[selected_model] # keeping only the TRUE predictors

```

::::: {layout="[0.25, 0.75]"}
<div>

```{r}
#| include: false
#| label: tbl-all-sub-set-log-reg
#| tbl-cap: "Exhaustive Model Selection: Selected Predictors Based on BIC"

selected_vars |> kable()
```

</div>

<div>

```{r}
#| include: false
#| label: fig-all-subset-log-reg
#| fig-cap: "Exhaustive Model Selection, B) Forward Model Selection"
plot(best_log_reg_all_sub, scale = "bic")

```

</div>
:::::

```{=html}
<!-- 
Method 2: Forward Selection
-->
```

```{r}
#| include: false
# forward selection

best_log_reg_fwd <- regsubsets(heart_disease ~ .,
                         method = "forward",
                         nvmax = 17,
                         data = heart_train)


model_summary <- summary(best_log_reg_fwd)
best_idx <- which.min(model_summary$bic) # find the index in that corresponds to the model with
                                          # the lowest BIC
selected_model <- model_summary$which[best_idx, ] # select the row best_idx in the logical matrix --> selected_model is a logical vector
selected_model |> kable() 

selected_vars <- names(selected_model)[selected_model] # keeping only the TRUE predictors
selected_vars |> kable()
```

::::: {layout="[0.25, 0.75]"}
<div>

```{r}
#| include: false
#| label: tbl-fwd-log-reg
#| tbl-cap: "Forward Model Selection: Selected Predictors Based on BIC"

selected_vars |> kable()
```

</div>

<div>

```{r}
#| include: false
#| label: fig-fwd-log-reg
#| fig-cap: "Forward Model Selection (BIC)"

plot(best_log_reg_fwd, scale = "bic")
```

</div>
:::::

```{=html}
<!-- 
Method 3: Backward Selection
-->
```

```{r}
#| include: false
# backward selection

best_log_reg_bwd <- regsubsets(heart_disease ~ .,
                         method = "backward",
                         nvmax = 17,
                         data = heart_train)

model_summary <- summary(best_log_reg_bwd)
best_idx <- which.min(model_summary$bic) # find the index in that corresponds to the model with
                                          # the lowest BIC
selected_model <- model_summary$which[best_idx, ] # select the row best_idx in the logical matrix --> selected_model is a logical vector
selected_model |> kable() 

selected_vars <- names(selected_model)[selected_model] # keeping only the TRUE predictors
selected_vars |> kable()
```

::::: {layout="[0.25, 0.75]"}
<div>

```{r}
#| include: false
#| label: tbl-bwd-log-reg
#| tbl-cap: "Bakward Model Selection: Selected Predictors Based on BIC"

selected_vars |> kable()
```

</div>

<div>

```{r}
#| include: false
#| label: fig-bwd-log-reg
#| fig-cap: "Backward Model Selection (BIC)"

plot(best_log_reg_bwd, scale = "bic")
```

</div>
:::::

::::: {layout="[0.25, 0.75]"}
<div>

```{r}
#| echo: false
#| label: tbl-selected-vars-log-reg
#| tbl-cap: "Selected Predictors Based on BIC (for exhaustive, forward, and backward selection technqiues)"

selected_vars |> kable()
```

</div>

<div>

```{r}
#| echo: false
#| fig-cap: "A) Exhaustive Model Selection"

# Exhaustive
plot(best_log_reg_all_sub, scale = "bic")

```

```{r}
#| echo: false
#| fig-cap: "B) Forward Model Selection"
#| # Forward
plot(best_log_reg_fwd, scale = "bic")

```

```{r}
#| echo: false
#| fig-cap: "C) Backward Model Selection"
# Backward
plot(best_log_reg_bwd, scale = "bic")
```

</div>
:::::

```{r}
#| include: false

# create logistic regression using all predictors
full_log_reg <- glm(
  heart_disease ~ .,
  data = disease_data,
  family = "binomial"
)

summary(full_log_reg)

```

**Validating Final Model Assumptions**

Given the results of our model selection, we fit a linear regression
using the selected predictors listed above:

```{r}
#| include: false
# selected logistic model
logistic_model <- glm(
  heart_disease ~  smoking + alcohol_drinking + stroke + mental_health + diff_walking + sex + age_category + race + diabetic + gen_health + asthma + kidney_disease + skin_cancer,
  data = heartdisease,
  family = "binomial"
)


summary(logistic_model)
```

$$
\begin{aligned}
log(Odds_{i}) = \beta_0 + \beta_1Smoking_i + \beta_2Alcohol_i + \beta_3Stroke_i + \\
\beta_3MentalHealth_i + \beta_4DiffWalking_i + \beta_5Sex_i + \\
\beta_6Over55_i + \beta_7Asian_i + \beta_8Black_i + \beta_9Hispanic_i + \\
\beta_{10}Diabetic_i + \beta_{11}HealthFair_i + \beta_{12}HealthGood_i + \\
\beta_{13}HealthExcellent_i + \beta_{14}Asthma_i + \beta_{15}KidneyDisease_i + \\
\beta_{16}SkinCancer_i
\end{aligned}
$$

Interestingly, the logistic regression yielded a p-value of 0.128 (\>
0.05) for $\beta_{MentalHealth}$, indicating that the predictor does not
significantly impact the log-odds of an individual having a heart
attack. Given this, we removed `mental_health` from the final model. All
other p-values indicated that our selected variables were significant (p
\< 0.05).

```{r}
#| include: false

# FINAL logistic model
final_log_reg <- glm(
  heart_disease ~  smoking + alcohol_drinking + stroke + diff_walking + sex + age_category + race + diabetic + gen_health + asthma + kidney_disease + skin_cancer,
  data = heart_train,
  family = "binomial"
)


summary(final_log_reg)
```

$$
\begin{aligned}
log(Odds_{i}) = \beta_0 + \beta_1Smoking_i + \beta_2Alcohol_i + \beta_3Stroke_i + \\
+ \beta_3DiffWalking_i + \beta_4Sex_i + \\
\beta_5Over55_i + \beta_6Asian_i + \beta_7Black_i + \beta_8Hispanic_i + \\
\beta_{9}Diabetic_i + \beta_{10}HealthFair_i + \beta_{11}HealthGood_i + \\
\beta_{12}HealthExcellent_i + \beta_{13}Asthma_i + \beta_{14}KidneyDisease_i + \\
\beta_{15}SkinCancer_i
\end{aligned}
$$

[Binary Predictor Assumption]{.underline}

Our outcome variable (heart disease, where 0 is absence of heart disease
and 1 is presence) is binary and thus can be properly modeled using
logistic regression. The data cleaning process ensured that all
observations of independent, upholding the independence assumption of
the regression.

[Linearity Assumption]{.underline}

The linearity of the log-odds of heart disease must also be assessed
with each of its continuous predictor variables. However, after removing
`mental_health` from our regression, the model no longer contains any
continuous variables. As such, this assumption is satisfied by default.

[Sample Size Assumption]{.underline}

91.4% of the 319,795 observations do not have heart disease, while 8.6%
of the individuals do have heart disease. The same split is true for the
training data set we used to fit the logistic regression. Given that
less than 10% of the data have heart disease, we should exercise some
caution in assessing the model's results. However, we will move forward
with the analysis.

```{r}
#| include: false

# Total Data Set
heartdisease |> filter(heart_disease == "0") |> nrow() #292422 (91.4%)

heartdisease |> filter(heart_disease == "1") |> nrow()
#27373 (8.6%)

# Training Data Set
heart_train |> filter(heart_disease == "0") |> nrow() #233981 (91.5%)

heart_train |> filter(heart_disease == "1") |> nrow()
#21855 (8.6%)

```

[No Outliers Assumption]{.underline}

To identify potential outliers, we plotted the residuals vs fitted values (@fig-resids-fitted-log-reg), which indicated that observations 115488, 65997, and 96834 were abnormal values. Similarly, observations 52951, 160975, and 222531 had unusually large Cook's distance (@fig-cooks-log-reg). All of these potential outliers are compiled in @tbl-outliers. To test for the effect of these outliers, we removed all six observations from the training data set and refit the logistic regression. The new regression yielded nearly identical values for all coefficients and the same p-values. As such, we concluded that these observations did not significantly impact the regression outcome and decided to move forward with the analysis including all data points.

Since the raw residual plot (@fig-resids-fitted-log-reg) is not easily interpretable for a logistic regression, we opted to use a binned residual plot (@fig-binned-resids-log-reg). From this plot, we can see that the majority of the data points fall within the gray lines, which indicate 2 standard errors from the average residual value.

** TBD INTERPRETATION **



```{r}
#| echo: false
#| label: tbl-outliers
#| tbl-cap: "Outliers"

# identifying outliers
heart_train[c(115488,65997,96834, 52951, 160975,222531), ] |> kable() |>
  kable_styling() |>
  row_spec(0, angle = 90) |>
  column_spec(which(names(heart_train) == "race"), width = "3em")



```


```{r}
#| include: false

# remove outliers & refit the regression
heart_train_no_outliers <- heart_train[-c(115488,65997,96834, 52951, 160975,222531), ]


log_reg_no_outliers <- glm(
  heart_disease ~  smoking + alcohol_drinking + stroke + diff_walking + sex + age_category + race + diabetic + gen_health + asthma + kidney_disease + skin_cancer,
  data = heart_train_no_outliers,
  family = "binomial"
)


summary(log_reg_no_outliers)

```



::::: {layout="[0.45, 0.45]"}
:::{#plot1}
```{r}
#| echo: false
#| label: fig-resids-fitted-log-reg

# Residuals vs Fitted
plot(final_log_reg, which = 1)

```
:::
:::{#plot2}
```{r}
#| echo: false
#| label: fig-cooks-log-reg
# Cook's distance
plot(final_log_reg, which = 4)
```


:::
::::

::::: {layout="[0.45, 0.45]"}
:::{#plot1}
```{r}
#| include: false

# Scale Location
plot(final_log_reg, which = 3)
```
:::
:::{#plot2}

```{r}
#| include: false

# Q-Q residuals
plot(final_log_reg, which = 2)
```

:::
::::


```{r}
#| echo: false
#| label: fig-binned-resids-log-reg
#| fig-width: 4
#| fig-height: 3.5

library(arm)
binnedplot(fitted(final_log_reg), 
           residuals(final_log_reg, type = "response"), 
           nclass = NULL, 
           xlab = "Expected Values", 
           ylab = "Average residual", 
           main = "Binned residual plot", 
           cex.main = 1,
           cex.pts = 0.8, 
           col.pts = 1, 
           col.int = "gray")

```

[No Multicollinearity]{.underline}

As with our assessment of the MLR for BMI, we opted to use Generalized
Variance Inflation Factors (GVIF) to determine potential
multicollinearity of our predictor variables. Calculating the VIFs for
all predictors revealed values below 2, indicating no concerning degrees
of collinearity. As such, no predictors were removed from the model.

```{r}
#| include: false

# ASSUMPTION 4: No Multicollinearity

# determine potential multicollinearity using Generalized Variance Inflation Factors (GVIF)

vif(final_log_reg)
# all GVIF values are below 2, indicating that there is no collinearity --> we do not need to remove or combine variables to account for correlation
# VIF indicates "how much variance of estimated regression coefficient is increased due 
# to collinearity with other predictors"
# VIF values of 1 indicate no correlation
# VIF >5-10 indicate high multicollinearity

```

```{r}
#| include: false

# ASSUMPTION 3: linear relationship between continuous predictor & log-odds outcome


```

Finally, the normality and homoscedasticity of the residuals are not
required assumptions for logistic regression and as such were not
assessed for this analysis.

