```{r}
#| include: false

# loading libraries
library(tidyverse)
library(tidyr)
library(kableExtra)
library(pROC)

# load data
disease_data <- heartdisease

```

## Results

### Factors Associated with BMI ....

#### Descriptive Statistics

[Demographics]{.underline}

insert text

[Behaviors]{.underline}

insert texttt

[Health]{.underline}

insert texttt

#### Regression Analysis

The results of our multiple linear regression are presented in @tbl-mlr-reg and @tbl-mlr-stat.

The results show that `physical_activity` is negatively associated with `log(bmi)` and is statistically significant (p = *insert*, 95% CI: [*insert*, *insert*]).

```{r}
#| include: false
## IMPORTANT PREDICTORS


# physical_activity1	-0.0341034	0.0012078	-28.2370395	0.0000000	-0.0364705	-0.0317362

# gen_health
  # gen_healthFair	0.0361928	0.0026152	13.8395346	0.0000000	0.0310671	0.0413185
  # gen_healthGood	0.0346531	0.0027031	12.8198035	0.0000000	0.0293551	0.0399511
  # gen_healthExcellent	-0.0578620	0.0029013	-19.9435303	0.0000000	-0.0635485	-0.0521755

# smoking1	-0.0163608	0.0010499	-15.5824408	0.0000000	-0.0184187	-0.0143029

#age_category
  # age_category2	0.0547660	0.0015474	35.3924778	0.0000000	0.0517331	0.0577988
  # age_category3	-0.0018443	0.0014969	-1.2320452	0.2179340	-0.0047782	0.0010897

```


*lower physical activity, poorer general health, and smoking behavior. It is also expected to vary across demographic characteristics such as age, reflecting differences in lifestyle patterns and accumulated health risks over time.*

**Main variables of interest: physical activity, gen_health, smoking, age**


```{r}
#| include: false

final_mlr <- lm(log(bmi) ~ smoking + alcohol_drinking + stroke + physical_health +
                   diff_walking + sex + age_category + race + diabetic + physical_activity +
                   gen_health + sleep_time + asthma + skin_cancer, 
                 data = heart_train)

mlr_summary <- summary(final_mlr)

```

```{r}
#| include: false

# MAKING MLR TABLE


## CREATING TABLE

mlr_table <- as.data.frame(mlr_summary$coefficients)

## ADDING CONFIDENCE INTERVALS

mlr_ci <- as.data.frame(confint(final_mlr))
mlr_ci$Variable <- rownames(mlr_ci)

mlr_table <- cbind(mlr_table, mlr_ci[,1:2])

colnames(mlr_table)[ncol(mlr_table)-1] <- "CI_Lower"
colnames(mlr_table)[ncol(mlr_table)] <- "CI_Upper"

```


```{r}
#| echo: false
#| label: tbl-mlr-reg
#| tbl-cap: "Multiple Linear Regression Coefficients. "

mlr_table %>%
  kable(col.names = c(
    "Coefficient",
    "Estimate",
    "Std. Error",
    "t value",
    "p-value",
    "Lower CI",
    "Upper CI"
  ))

## Coefficients represent approximate percentage changes in BMI

```

```{r}
#| include: false

# TABLE FOR MLR STATS

## GETTING STATS

n <- nobs(final_mlr)
r2 <- (0.145379)
adj_r2 <- (0.145328)

mlr_stats <- data.frame(
  Statistic = c("R-squared", "Adjusted R-squared", "N"),
  Value = c(
    mlr_model_sum$r.squared,
    mlr_model_sum$adj.r.squared,
    nobs(final_mlr)
  )
)

```


```{r}
#| echo: false
#| label: tbl-mlr-stat
#| tbl-cap: "Multiple Linear Regression Statistics"

mlr_stats %>% 
  kable(col.names = c("Statistic", "Value"))

```



### Predicting Heart Attack Risk ....

#### Descriptive Statistics

**TBD FINAL REGRESSION EQUATION**

**TBD PROBABILITY & ODDS RATIO INTERPRETATION**
The ultimate result of our logistic regression on the odds of heart disease are captured in @tbl-log-reg-coeffs.
```{r}
#| echo: false
#| label: tbl-log-reg-coeffs
#| tbl-cap: "Logistic Regression Coefficients. Red rows indicate coefficients that increase the odds of heart disease. Green rows indicate coefficients that decrease the odds of heart disease. Gray rows indicate coefficients with insignificant p-values."
reg_sum <- summary(final_log_reg)

coefficients <- as.data.frame(reg_sum$coefficients)
coeff_sorted <- coefficients[order(coefficients$Estimate, decreasing = TRUE), ]


# calculate the odds & probability
coeff_sorted <- coeff_sorted |>
  mutate(
    odds_coeff = exp(coeff_sorted$Estimate),
    #prob = odds_coeff / (1 + odds_coeff),
    ci_low = exp(Estimate - 2*`Std. Error`),
    ci_high = exp(Estimate + 2*`Std. Error`)
  )

colnames(coeff_sorted) <- c("Estimate (log-odds)", "Std. Error", "Z value", "P-value", "Estimate (odds)",  "CI low (odds)", "CI high (odds)")

coeff_sorted |> kable() |>
  kable_styling(full_width = FALSE) |>
  column_spec(1:ncol(coeff_sorted), width = "4em") |>
  row_spec(which(coeff_sorted$`Estimate (log-odds)` >= 0),  background = "#ffe6e6") |>
  row_spec(which(coeff_sorted$`Estimate (log-odds)` < 0), background = "#e6ffe6") |>
  row_spec(which(coeff_sorted$`P-value` > 0.05), background = "#c7d1d9")
```

```{r}
#| include: false

# create modified data set with categorical sleep, age, & bmi variables

young = c("18-24", "25-29", "30-34", "35-39")
middle_age = c("40-44", "45-49", "50-54", "55-59")
old = c("60-64", "65-69", "70-74", "75-79", "80 or older")

disease2 <- disease_data |>
  mutate(sleep_time = case_when(
    sleep_time <= 10 & sleep_time >= 7 ~ "Healthy Sleep (7-10 hrs)",
    sleep_time < 7 ~ "Sleep Deprived (<7 hrs)",
    sleep_time > 10 ~ "Extra Sleep (>10 hrs)",
    TRUE ~ NA_character_),
    
    bmi = case_when(
    bmi < 19 ~ "Underweight",
    bmi >= 19 & bmi < 25 ~ "Normal Weight",
    bmi >= 25 & bmi < 30 ~ "Overweight",
    bmi >= 30 ~ "Obese",
    TRUE ~ NA_character_),
    
   # age_category = case_when(
 #   age_category %in% young ~ "Young (18-39)",
  #  age_category %in% middle_age ~ "Middle Aged (40-59)",
  #  age_category %in% old ~ "Elderly (60-80+)",
  #  TRUE ~ NA_character_)
    
    )
```

```{r}
#| include: false

# testing for correlation

#cor(disease_data$smoking, disease_data$asthma) #0.02414853
#cor(disease_data$diabetic, disease_data$heart_disease) #0.1747823

```

[Demographics]{.underline}

Compared to White peers, and holding all other variables constant, our model predicts with that Hispanic, Black, and Asian individuals have 0.75 (95% CI: \[0.70, 0.81\]), 0.70 (95% CI: \[0.66, 0.76\]), and 0.60 (95% CI: \[0.51, 0.71\]) times the odds of having heart disease, respectively (@fig-heartdisease-race; @tbl-log-reg-coeffs) . In short, White individuals are predicted to have the highest likelihood of heart attack.

As one may expect, the predicted risk for heart disease increases with age. Compared to 18-34 year olds of the same sex, race, and health conditions, 35-54 year olds are expected to have 2.97 the odds of heart disease (95% CI: [2.61, 3.37]). This risk increases even more dramatically for ages 55 years and older, whose odds of heart diseae are predicted to be 8.9 times (95% CI: [7.89, 10.06]) that of 18-34 year olds (@fig-heartdisease-age; @tbl-log-reg-coeffs).

Sex also has a significant impact on the odds of having heart disease. Males have 0.55 times (95% CI: [0.53, 0.57]) the odds of have heart disease compared to females, holding all other predictors constant (@fig-heartdisease-sex; @tbl-log-reg-coeffs).

:::::: {layout="[0.3, 0.3, 0.3]"}
::: {#plot1}
```{r}
#| echo: false
#| label: fig-heartdisease-race
#| fig-cap: "Race"
## DEMOGRAPHICS

# RACE
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = race)) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Race") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  #scale_fill_manual(values = c(1 = "tomato", 0 = "skyblue")) +
  theme_classic() + 
  theme(legend.position = "bottom")
```
:::

::: {#plot2}
```{r}
#| echo: false
#| label: fig-heartdisease-sex
#| fig-cap: "Sex"
# SEX
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = factor(sex))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Sex") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  #scale_fill_manual(values = c(1 = "tomato", 0 = "skyblue")) +
  theme_classic() + 
  theme(legend.position = "bottom")
```
:::

::: {#plot3}
```{r}
#| echo: false
#| label: fig-heartdisease-age
#| fig-cap: "Age"

# AGE
ggplot(data = disease2, aes(x = factor(heart_disease), fill = age_category)) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease",
       y = "Count",
       fill = "Age Group") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("1" = "orange", "2" = "tomato", "3" = "skyblue"),
    labels = c("18-34", "35-54", "55 and older")) +
  theme_classic() + 
  theme(legend.position = "bottom")




```
:::
::::::
```{r}
#| include: false
# AGE
ggplot(data = disease_data, aes(x = age_category, fill = factor(heart_disease))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Age",
       y = "Count",
       fill = "Heart Disease Status") +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  scale_x_discrete(labels = c("1" = "18-34",
                              "2" = "35-54",
                              "3" = "55 and older")) +
  theme_classic() + 
  theme(legend.position = "bottom")
```

:::
::::


[Behavior]{.underline}


As we expected, our model predicts that individuals who smoke are at higher risk for heart disease. In particular, their odds of having a heart attack are 1.41 times (95% CI: [1.35, 1.46]) the odds for individuals who do not smoke, holding all other predictors constant (@fig-heartdisease-smoking; @tbl-log-reg-coeffs). Interestingly, alcohol consumption is predicted to have a significant negative effect on the risk of heart disease. Individuals who drink alcohol have 0.70 times (95% CI: [0.64, 0.76]) the odds of heart disease compared to individuals who do not drink alcohol (@fig-heartdisease-alcohol; @tbl-log-reg-coeffs).

:::: {layout="[0.45, 0.45]"}
:::{#plot1}
```{r}
#| echo: false
#| label: fig-heartdisease-smoking
#| fig-cap: "Smoking"

# SMOKING
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = factor(smoking))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Smoker") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() + 
  theme(legend.position = "bottom")



```
:::
:::{#plot2}
```{r}
#| echo: false
#| label: fig-heartdisease-alcohol
#| fig-cap: "Alcohol Consumption"

# ALCOHOL
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = factor(alcohol_drinking))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Drinks Alcohol") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() + 
  theme(legend.position = "bottom")
```
:::
::::


[Medical Conditions]{.underline}

Overall, individuals categorized as having 'Fair', 'Good', and 'Excellent' general health were predicted to have a lower risk of heart disease. Compared to the odds of heart disease for those with 'Poor' general health, individuals in the other health categories had 0.68 (95% CI: [0.64, 0.72]), 0.42 (95% CI: [0.40, 0.45]), and 0.13 (95% CI: [0.12, 0.14]) times the odds of heart attack, respectively (@fig-heartdisease-gen-health; @tbl-log-reg-coeffs).



```{r}
#| echo: false
#| label: fig-heartdisease-gen-health

# GENERAL HEALTH
disease_data |>
  filter(!is.na(gen_health)) |>
  ggplot(aes(x = factor(heart_disease), fill = factor(gen_health))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "General Health") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  #scale_fill_manual(
 #   values = c("0" = "skyblue", "1" = "tomato"),
#    labels = c("No", "Yes")
#  ) +
  theme_classic() + 
  theme(legend.position = "bottom")
```

On the other hand, pre-existing health complications such as diabetes, kidney disease, skin cancer, asthma, stroke, and difficulty walking significantly increased the odds of heart disease. Diabetic patients have 1.61 times (95% CI: [1.55, 1.68]) the odds, patients with skin cancer have 1.33 times (95% CI: [1.27, 1.40]) the odds, patients with kidney disease have 1.87 times (95% CI: [1.76, 1.98]) the odds, patients with asthma have 1.25 times (95% CI: [1.19, 1.31]) the odds,  patients who have had a stroke have 2.96 times (95% CI: [2.81, 3.13]) the odds, and patients with difficulty walking have 1.32 (95% CI: [1.27, 1.38]) the odds of heart attack compared to individuals without those respective health conditions, holding all other predictors constant (@fig-heartdisease-diabetes; @fig-heartdisease-kidney; @fig-heartdisease-cancer; @fig-heartdisease-asthma; @fig-heartdisease-stroke @tbl-log-reg-coeffs). Overall, having a stroke increases an individuals risk for heart attack to the highest degree of the health conditions mentioned. 

:::: {layout="[0.3, 0.3, 0.3]"}
:::{#plot1}
```{r}
#| echo: false
#| label: fig-heartdisease-diabetes
#| fig-cap: "Diabetes"

## MEDICAL CONDITIONS

# DIABETES
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = factor(diabetic))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Diabetic") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() + 
  theme(legend.position = "bottom")
```
:::


:::{#plot2}
```{r}
#| echo: false
#| label: fig-heartdisease-kidney
#| fig-cap: "Kidney Disease"

# KIDNEY DISEASE
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = factor(kidney_disease))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Kidney Disease") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() + 
  theme(legend.position = "bottom")
```
:::

:::{#plot3}

```{r}
#| echo: false
#| label: fig-heartdisease-asthma
#| fig-cap: "Asthma"
# ASTHMA
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = factor(asthma))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Asthma") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() + 
  theme(legend.position = "bottom")
```
:::
::::


:::: {layout="[0.3, 0.3, 0.3]"}

:::{#plot2}

```{r}
#| echo: false
#| label: fig-heartdisease-cancer
#| fig-cap: "Skin Cancer"
# SKIN CANCER
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = factor(skin_cancer))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Skin Cancer") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() + 
  theme(legend.position = "bottom")

```
:::

:::{#plot3}
```{r}
#| echo: false
#| label: fig-heartdisease-stroke
#| fig-cap: "Stroke"
# STROKE
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = factor(stroke))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Stroke") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() + 
  theme(legend.position = "bottom")
```
:::

:::{#plot1}
```{r}
#| echo: false
#| label: fig-heartdisease-walking
#| fig-cap: "Difficulty Walking"
# DIFFICULTY WALKING
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = factor(diff_walking))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Difficulty Walking") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() + 
  theme(legend.position = "bottom")
```

:::

::::


#### Regression Analysis



```{r}
#| include: false


# calculate the odds & probability
heart_test <- heart_test |>
  mutate(
    log_odds = predict(final_log_reg, newdata = heart_test, type = "link"),
    odds = exp(log_odds),
    prob = odds / (1 + odds)
  )

```


To assess the sensitivity and specificity of our logistic regression model and to determine an appropriate probability threshold for classification, we plotted a Receive Operating Characteristic (ROC) curve (@fig-roc). The Area Under the Curve (AUC) is 0.83, indicating that the model's classification of individuals with and without heart disease is much better than random (AUC = 0.5, represented by the y = x dashed line). We opted to use Youden's J statistic  to determine a probability threshold. This statistic identifies the maximum vertical distance between the ROC curve and the $y = x$ line, thereby maximizing the difference between the true positive rate (sensitivity) and the false positive rate (1 - specificity). We obtained a Youden J statistic of 0.09886 and set this as our threshold for subsequent analysis.
```{r}
#| echo: false
#| label: fig-roc
#| fig-cap: "Receiver Operating Characteristic (ROC) Curve"
#| fig-width: 3.5
#| fig-height: 3.5

# calculate predicted probabilities
prob <- predict(final_log_reg, newdata = heart_test, type = "response")

roc <- roc(heart_test$heart_disease, prob)
df <- data.frame(sensitivity = roc$sensitivities, specificity = roc$specificities)

ggplot(data = df, aes(x = 1 - specificity, y = sensitivity)) +
  geom_point() +
  labs(
    x = "1 - Specificity (False Positive Rate)",
    y = "Sensitivity (True Positive Rate)") + 
  geom_abline(slope = 1,
              intercept = 0, 
              linetype = "dashed", 
              color = "steelblue") +
  theme_classic()


# calculate the Area Under the Curve
auc(roc) # 0.8292

# pick a threshold
# Method 1: Youden's J statistic (maximum vertical distance between ROC curve & diagonal line)
  # maximises the diff between true positive rate & false positive rate
youden <- coords(roc, "best", best.method = "youden")
  # threshold = 0.09886

# Method 2: threshold closest to (0, 1) (perfect classification)
topleft <- coords(roc, "best", best.method = "closest.topleft")
  # threshold = 0.1161

```


Using this threshold, we constructed a confusion matrix for the predicted heart disease outcomes on our test data (see section 2 for details on how we developed the train and test data) (@tbl-confusion-matrix). From this confusion matrix, we then calculated the rates for true positive, false positive, true negative, and false negative detection (@tbl-confusion-matrix-rates). At the threshold of 0.09886, our model can accurately identify 80.6% of individuals with heart disease, indicating high model sensitivity. The model's ability to accurately classify individuals as not having heart disease is also quite high at 70.2%, indicating high model specificity. This being said, the model will still incorrectly diagnose an individual has having heart disease 29.8% of the time and fails to properly identify a person as having heart disease 19.4% of the time.

::::: {layout="[0.7, 0.3]"}
<div>

```{r}
#| echo: false
#| label: tbl-confusion-matrix
#| tbl-cap: "Confusion Matrix for Predicted Heart Disease Outcomes on Test Data"


# classify heart disease based on probability
threshold <- youden$threshold
pred_class <- ifelse(prob > threshold, 1, 0)

# create confusion matrix
confusion_matrix <- table(Predicted = pred_class, Actual = heart_test$heart_disease) 

rownames(confusion_matrix) <- c("No Heart Disease (Pred)", "Heart Disease (Pred)")
colnames(confusion_matrix) <- c("No Heart Disease (Actual)", "Heart Disease (Actual)")

confusion_matrix|> kable()
```

</div>

<div>

```{r}
#| echo: false
#| label: tbl-confusion-matrix-rates

# calculate true positive, false positive, false negative, and true negative rates
TN <- confusion_matrix[1, 1]
FN <- confusion_matrix[1, 2]
FP <- confusion_matrix[2, 1]
TP <- confusion_matrix[2, 2]

# measures of classification error
FNR <- FN/(TP + FN)
FPR <- FP/(TN + FP)

# measures of classification ability
TPR <- TP/(TP + FN) # sensitivity
TNR <- TN/(TN + FP) # specificity

# make a table
rates_tbl <- data.frame(
  metric = c("True Positive Rate", 
             "False Positive Rate",
             "False Negative Rate",
             "True Negative Rate"),
  value = c(TPR, FPR, FNR, TNR)
)

rates_tbl |> kable()

```

</div>
:::::

```{=html}
<!-- 
If you have more than 2 research questions, try to group them into no 
more than 2 groups. Rename the subsections to give a short description of the
group of questions (e.g. "Landscape of US colleges" and "Factors associated with dropout")
-->
```
