```{r}
#| include: false

# loading libraries
library(tidyverse)
library(tidyr)
library(kableExtra)

# load data
disease_data <- readRDS("../disease_data/disease-prepared.RDS")
```

## Results

### Factors Associated with BMI ....

#### Descriptive Statistics

#### Regression Analysis

### Predicting Heart Attack Risk ....

#### Descriptive Statistics

```{r}
#| include: false

# create modified data set with categorical sleep, age, & bmi variables

young = c("18-24", "25-29", "30-34", "35-39")
middle_age = c("40-44", "45-49", "50-54", "55-59")
old = c("60-64", "65-69", "70-74", "75-79", "80 or older")

disease2 <- disease_data |>
  mutate(sleep_time = case_when(
    sleep_time <= 10 & sleep_time >= 7 ~ "Healthy Sleep (7-10 hrs)",
    sleep_time < 7 ~ "Sleep Deprived (<7 hrs)",
    sleep_time > 10 ~ "Extra Sleep (>10 hrs)",
    TRUE ~ NA_character_),
    
    bmi = case_when(
    bmi < 19 ~ "Underweight",
    bmi >= 19 & bmi < 25 ~ "Normal Weight",
    bmi >= 25 & bmi < 30 ~ "Overweight",
    bmi >= 30 ~ "Obese",
    TRUE ~ NA_character_),
    
    age_category = case_when(
    age_category %in% young ~ "Young (18-39)",
    age_category %in% middle_age ~ "Middle Aged (40-59)",
    age_category %in% old ~ "Elderly (60-80+)",
    TRUE ~ NA_character_)
    
    )
```

```{r}
#| include: false

# testing for correlation

cor(disease_data$smoking, disease_data$asthma) #0.02414853
cor(disease_data$diabetic, disease_data$heart_disease) #0.1747823

```

:::::: {layout="[0.3, 0.3, 0.3]"}
::: {#plot1}
```{r}
#| echo: false
## DEMOGRAPHICS

# RACE
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = race)) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Sex") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  #scale_fill_manual(values = c(1 = "tomato", 0 = "skyblue")) +
  theme_classic() + 
  theme(legend.position = "bottom")
```
:::

::: {#plot2}
```{r}
#| echo: false
# SEX
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = factor(sex))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Sex") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  #scale_fill_manual(values = c(1 = "tomato", 0 = "skyblue")) +
  theme_classic() + 
  theme(legend.position = "bottom")
```
:::

::: {#plot3}
```{r}
#| echo: false

# AGE
ggplot(data = disease2, aes(x = factor(heart_disease), fill = age_category)) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease",
       y = "Count",
       fill = "Age Group") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  #scale_fill_manual(values = c(1 = "tomato", 0 = "skyblue")) +
  theme_classic() + 
  theme(legend.position = "bottom")



ggplot(data = disease_data, aes(x = age_category, fill = factor(heart_disease))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Age",
       y = "Count",
       fill = "Heart Disease Status") +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  theme_classic() + 
  theme(legend.position = "bottom")


```
:::
::::::

:::::: {layout="[0.3, 0.3, 0.3]"}
<div>

</div>

<div>

</div>

<div>

</div>
::::::

::::: {layout="[0.45, 0.45]"}
::: column
```{r}
#| echo: false
## BODY MASS INDEX (BMI)

ggplot(data = disease_data, aes(x = bmi, fill = factor(heart_disease))) +
  geom_density(alpha = 0.5, color = "black") + 
  labs(x = "BMI",
       y = "Count",
       fill = "Heart Disease") +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() +
  theme(legend.position = "bottom")

# BMI
ggplot(data = disease_data, aes(y = bmi, x = factor(heart_disease), 
                            fill = factor(heart_disease))) +
  geom_boxplot() + 
  labs(x = "Heart Disease", y = "BMI") + 
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() +
  theme(legend.position = "none")
```
:::

::: column
```{r}
#| echo: false

# BMI
ggplot(data = disease2, aes(x = factor(heart_disease), fill = bmi)) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease",
       y = "Count",
       fill = "BMI") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  theme_classic() + 
  theme(legend.position = "bottom")
```
:::
:::::

```{r}
#| include: false

# BMI
ggplot(data = disease_data, aes(y = bmi, x = factor(heart_disease), 
                            fill = factor(heart_disease))) +
  geom_boxplot() + 
  labs(x = "Heart Disease", y = "BMI") + 
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() +
  theme(legend.position = "none")

```

:::::: {layout="[0.3, 0.3, 0.3]"}
<div>

```{r}
#| echo: false
## BEHAVIOR

# SLEEP
ggplot(data = disease2, aes(x = factor(heart_disease), fill = sleep_time)) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease",
       y = "Count",
       fill = "Sleep") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  theme_classic() + 
  theme(legend.position = "bottom")
```

</div>

<div>

```{r}
#| echo: false

# ALCOHOL
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = factor(alcohol_drinking))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Drinks Alcohol") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() + 
  theme(legend.position = "bottom")
```

</div>

<div>

```{r}
#| echo: false

# SMOKING
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = factor(smoking))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Smoker") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() + 
  theme(legend.position = "bottom")


# PHYSICAL ACTIVITY
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = factor(physical_activity))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Physically Active") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() + 
  theme(legend.position = "bottom")
```

</div>
::::::

:::::: {layout="[0.3, 0.3, 0.3]"}
<div>

```{r}
#| echo: false

## MEDICAL CONDITIONS

# DIABETES
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = factor(diabetic))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Diabetic") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() + 
  theme(legend.position = "bottom")
```

</div>

<div>

```{r}
#| echo: false

# KIDNEY DISEASE
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = factor(kidney_disease))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Kidney Disease") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() + 
  theme(legend.position = "bottom")
```

</div>

<div>

```{r}
#| echo: false
# ASTHMA
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = factor(asthma))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Asthma") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() + 
  theme(legend.position = "bottom")
```

</div>
::::::

:::::: {layout="[0.3, 0.3, 0.3]"}
<div>

```{r}
#| echo: false
# SKIN CANCER
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = factor(skin_cancer))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Skin Cancer") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() + 
  theme(legend.position = "bottom")

```

</div>

<div>

```{r}
#| echo: false
# STROKE
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = factor(stroke))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Stroke") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() + 
  theme(legend.position = "bottom")
```

</div>

<div>

</div>
::::::

#### Regression Analysis

-   final regression equation

-   interpretation of coefficients

-   odds interpretation (odds ratio, confidence interval)

-   probability interpretation

-   confusion matrix (sensitivity & specificity)

-   Receiver operating characteristic (ROC) curve

```{=html}
<!-- 
If you have more than 2 research questions, try to group them into no 
more than 2 groups. Rename the subsections to give a short description of the
group of questions (e.g. "Landscape of US colleges" and "Factors associated with dropout")
-->
```
