```{r}
#| include: false

# loading libraries
library(tidyverse)
library(tidyr)
library(kableExtra)
library(pROC)
library(broom)    # For tidying model objects
library(modelr)   # For generating prediction grids
library(scales)   # For better axis labels
library(yardstick) # For ROC curve

# load data
heartdisease <- readRDS("../disease_data/heart-prepared.RDS")
disease_data <- heartdisease

```

## Results

### Factors Associated with BMI

#### Descriptive Statistics & Model Results

```{r}
#| include: false

## FROM TRAIN DATA

test_finalmlr <- lm(log(bmi) ~ smoking + alcohol_drinking + stroke + physical_health +
                   diff_walking + sex + age_category + race + diabetic + physical_activity +
                   gen_health + sleep_time + asthma + skin_cancer, 
                 data = heart_train)

test_mlr_summary <- summary(test_finalmlr)


## CREATING TABLE

test_mlr_table <- as.data.frame(test_mlr_summary$coefficients)

## ADDING CONFIDENCE INTERVALS

test_mlr_ci <- as.data.frame(confint(test_finalmlr))
test_mlr_ci$Variable <- rownames(test_mlr_ci)

test_mlr_table <- cbind(test_mlr_table, test_mlr_ci[,1:2])

colnames(test_mlr_table)[ncol(test_mlr_table)-1] <- "CI_Lower"
colnames(test_mlr_table)[ncol(test_mlr_table)] <- "CI_Upper"

test_mlr_table %>%
  kable(col.names = c(
    "Coefficient",
    "Estimate",
    "Std. Error",
    "t value",
    "p-value",
    "Lower CI",
    "Upper CI"
  ))

```


```{r}
#| include: false

## TRAIN DATA RESULTS

# smoking1	-0.0163608	0.0010499	-15.5824408	0.0000000	-0.0184187	-0.0143029

# gen_health
  # gen_healthFair	0.0361928	0.0026152	13.8395346	0.0000000	0.0310671	0.0413185
  # gen_healthGood	0.0346531	0.0027031	12.8198035	0.0000000	0.0293551	0.0399511
  # gen_healthExcellent	-0.0578620	0.0029013	-19.9435303	0.0000000	-0.0635485	-0.0521755

# physical_activity1	-0.0341034	0.0012078	-28.2370395	0.0000000	-0.0364705	-0.0317362

# age_category
  # age_category2	0.0547660	0.0015474	35.3924778	0.0000000	0.0517331	0.0577988
  # age_category3	-0.0018443	0.0014969	-1.2320452	0.2179340	-0.0047782	0.0010897
```


[Demographics: Age]{.underline}

The distribution of log(BMI) varies across age groups (@fig-logbmi-age). Individuals who are 35-54 years old appear to have a higher median log(BMI) compared to those who are 18-34 years old. Those who are 55+ years old have a slightly lower median (BMI) than 35-54 year olds that is also slightly higher than for 18-34 year olds. Overall, BMI appears to increase from younger to middle-aged adults.


```{r}
#| echo: false
#| label: fig-logbmi-age
#| fig-width: 4
#| fig-height: 3

heart_train %>%
  filter(!is.na(age_category), !is.na(bmi)) %>%
  ggplot() +
  aes(x = age_category, y = log(bmi), fill = age_category) +
  geom_boxplot() +
  labs(x = "Age Groups", y = "log(BMI)", fill = "Age Groups") +
  theme_classic() +
  scale_fill_manual(values = c(
    "1" = "orange",
    "2" = "tomato",
    "3" = "skyblue"),
    labels = c(
      "1" = "18-34",
      "2" = "35-54",
      "3" = "55 and older")
  ) +
  scale_x_discrete(labels = c(
  "1" = "18-34",
  "2" = "35-54",
  "3" = "55 and older"
))

```


[Behaviors]{.underline}

Physical activity is also associated with differences in BMI. Individuals who report engaging in physical activity tend to have lower log(BMI) compared to those who do not, as seen by a lower median and distribution (@fig-logbmi-physical).

Smoking, on the other hand, shows a less pronounced relationship with BMI. While smokers appear to have a slightly lower median log(BMI) than non-smokers, the difference between the groups is relatively small (@fig-logbmi-smoke).

::::: {layout="[0.45, 0.45]"}
::: {#plot1}
```{r}
#| echo: false
#| label: fig-logbmi-physical

heart_train %>% 
  filter(!is.na(physical_activity), !is.na(bmi)) %>%
  ggplot() +
  aes(x = physical_activity, y = log(bmi), fill = physical_activity) %>%
  geom_boxplot() +
  labs(x = "Physical Activity", y = "log(BMI)", fill = "Physical Activity") +
  theme_classic() +
  scale_fill_manual(values = c(
    "0" = "tomato",
    "1" = "skyblue"),
    labels = c(
      "0" = "No",
      "1" = "Yes")
  ) +
  scale_x_discrete(labels = c(
  "0" = "No",
  "1" = "Yes"
))
  
```
:::

:::{#plot2}
```{r}
#| echo: false
#| label: fig-logbmi-smoke

heart_train %>% 
  filter(!is.na(smoking), !is.na(bmi)) %>%
  ggplot() +
  aes(x = smoking, y = log(bmi), fill = smoking) +
  geom_boxplot() +
  labs(x = "Smoking Status", y = "log(BMI)", fill = "Smoking Status") +
  theme_classic() +
    scale_fill_manual(values = c(
    "0" = "tomato",
    "1" = "skyblue"),
    labels = c(
      "0" = "No",
      "1" = "Yes")) +
  scale_x_discrete(labels = c(
  "0" = "No",
  "1" = "Yes"))
```
:::
::::

[Health]{.underline}

General health status shows a clear relationship with BMI. Individuals reporting excellent health tend to have lower BMI, while those reporting fair or poor health tend to have higher BMI. This suggests a gradient where better self-reported health is associated with lower BMI (@fig-logbmi-genhealth).


```{r}
#| echo: false
#| label: fig-logbmi-genhealth

heart_train %>% 
  filter(!is.na(gen_health), !is.na(bmi)) %>%
  ggplot() +
  aes(x = gen_health, y = log(bmi), fill = gen_health) +
  geom_boxplot() +
  labs(x = "General Health", y = "log(BMI)", fill = "General Health") +
  theme_classic() +
  scale_fill_manual(values = c(
    "Poor" = "tomato",
    "Fair" = "orange",
    "Good" = "gold",
    "Excellent" = "#228B22"))

```


[Combined Effects of Behavior and Health]{.underline}

To further explore whether the relationship between physical activity and BMI differs depending on health, we also examined the interaction between physical activity and general health (@fig-logbmi-physhealth). Across all levels of general health, individuals who report engaging in physical activity consistently had lower log(BMI) compared to those who did not. However, the magnitude of this difference appears relatively constant across all self-reported health categories, suggesting there may not be a strong interaction effect between physical activity and general health on log(BMI).

```{r}
#| echo: false
#| label: fig-logbmi-physhealth

heart_train %>%
  filter(!is.na(gen_health), !is.na(bmi), !is.na(physical_activity)) %>%
  ggplot(aes(
    x = factor(physical_activity, levels = c(0,1), labels = c("No","Yes")),
    y = log(bmi),
    fill = factor(physical_activity, levels = c(0,1), labels = c("No","Yes"))
  )) +
  geom_boxplot() +
  facet_wrap(~ gen_health) +
  labs(
    x = "Physical Activity",
    y = "log(BMI)",
    fill = "Physical Activity"
  ) +
  theme_classic() +
  scale_fill_manual(values = c("No" = "tomato", "Yes" = "skyblue"))

```


The patterns observed in the above descriptive analyses are further evaluated using a multiple linear regression model. The results of the regression on log(BMI) are captured in @fig-mlr-coefficients. Most predictors included in the final model significantly (p < 0.05) contributed to variation in log(BMI), with the exception of age category 55+, which was not statistically significant after adjusting for other variables.

```{r}
#| echo: false
#| label: fig-mlr-coefficients
#| fig-cap: "Coefficients of multiple linear regression model. Blue data points represent the estimated effect of each predictor on log(BMI). The red dashed line represents no effect on log(BMI). Bars represent 95% confidence intervals. Most effects are significant (p < 0.05), aside from age category 55+."

library(broom)

mlr_tidy <- tidy(final_mlr, conf.int = TRUE) |>
  filter(term != "(Intercept)") |>
  mutate(
    term = case_match(term,
      "smoking1" ~ "Smoker",
      "alcohol_drinking1" ~ "Drinks Alcohol",
      "stroke1" ~ "Stroke",
      "physical_health" ~ "Physical Health",
      "diff_walking1" ~ "Difficulty Walking",
      "sex1" ~ "Male",
      "age_category2" ~ "35–54 years old",
      "age_category3" ~ "55 and older",
      "raceAsian" ~ "Asian",
      "raceBlack" ~ "Black",
      "raceAmerican Indian/Alaskan Native" ~ "Native American",
      "diabetic1" ~ "Diabetic",
      "gen_healthFair" ~ "Fair Health",
      "gen_healthGood" ~ "Good Health",
      "gen_healthExcellent" ~ "Excellent Health",
      "sleep_time" ~ "Sleep Time",
      "asthma1" ~ "Asthma",
      "skin_cancer1" ~ "Skin Cancer",
      .default = term
    ),
    term = reorder(term, estimate)
  )

ggplot(mlr_tidy, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high), width = 0.2) +
  geom_point(size = 2, color = "steelblue") +
  labs(
    x = "Effect on log(BMI) (with 95% confidence interval)",
    y = NULL
  ) +
  theme_classic()
```




Moving forward, these observed patterns are further evaluated using multiple linear regression to assess whether these relationships persist after adjusting for other variables. Notably, `physical_activity` and `general_health` show the most visually distinct differences in BMI across groups, suggesting they are likely to remain among the strongest predictors in the regression model. In contrast, `smoking` and `age_category` display more modest differences, indicating that their relationships with BMI may be weaker or influenced by other factors after adjustment.


#### Regression Analysis

The results of our multiple linear regression are presented in @tbl-mlr-reg.

Physical activity is negatively associated with `log(bmi)` and is statistically significant. Holding all other factors constant, individuals who report engaging in physical activity have approximately 3.42% lower BMI compared to those who do not (p < 0.05, 95% CI: [-0.039, -0.030]). This aligns with expectations and supports the idea that engaging in physical activity is an important factor associated with lower BMI. This suggests that consistent exercise helps manage weight and burn calories, which makes maintaining a lower body mass much easier in the long run.

General health is also significantly associated with `log(bmi)`. Compared to individuals reporting poor health, those reporting fair health have approximately 3.81% higher BMI (p < 0.05, 95% CI: [0.028, 0.048]), those reporting good health have 3.31% higher BMI (p < 0.05, 95% CI: [0.023, 0.044]), and those reporting excellent health have 5.77% lower BMI (p < 0.05, 95% CI: [-0.069, -0.046]). This gradient across health categories reinforces the strength and consistency of this relationship and suggests that self-reported health status is closely linked to BMI.

Smoking is negatively associated with `log(bmi`). Holding all other factors constant, individuals who smoke have about 1.42% lower BMI compared to those who do not (p < 0.05, 95% CI: [-0.018, -0.010]). Although statistically significant, the magnitude of this effect is relatively small, suggesting that smoking may not be a practically important predictor of BMI despite its statistical significance, particularly given the large sample size.

Lastly, age is also significantly associated with `log(bmi)`, but only for individuals aged 35-54. Compared to individuals aged 18–34, those aged 35–54 have approximately 5.74% higher BMI (p < 0.05, 95% CI: [0.051, 0.063]). There is no statistically significant difference for individuals aged 55 and older. This suggests that BMI differences may be more pronounced in middle adulthood, while differences between younger and older adults are less distinct after accounting for other factors. This is slightly different from the patterns observed in the descriptive statistics, suggesting that differences in BMI by age may be explained by other variables included in the model.


```{r}
#| include: false
## IMPORTANT PREDICTORS


# physical_activity1	-0.0342309	0.0024034	-14.2429058	0.0000000	-0.0389416	-0.0295203

# gen_health
  # gen_healthFair	0.0380583	0.0052157	7.2968967	0.0000000	0.0278355	0.0482812
  # gen_healthGood	0.0330801	0.0053876	6.1400901	0.0000000	0.0225204	0.0436398
  # gen_healthExcellent	-0.0577355	0.0057809	-9.9873372	0.0000000	-0.0690662	-0.0464049

# smoking1	-0.0142997	0.0020833	-6.8639016	0.0000000	-0.0183831	-0.0102164

#age_category
  # age_category2	0.0574385	0.0030709	18.7038451	0.0000000	0.0514193	0.0634576
  # age_category3	0.0004762	0.0029635	0.1606875	0.8723403	-0.0053323	0.0062846

```


```{r}
#| include: false

final_mlr <- lm(log(bmi) ~ smoking + alcohol_drinking + stroke + physical_health +
                   diff_walking + sex + age_category + race + diabetic + physical_activity +
                   gen_health + sleep_time + asthma + skin_cancer, 
                 data = heart_test)

mlr_summary <- summary(final_mlr)

```

```{r}
#| include: false

# MAKING MLR TABLE


## CREATING TABLE

mlr_table <- as.data.frame(mlr_summary$coefficients)

## ADDING CONFIDENCE INTERVALS

mlr_ci <- as.data.frame(confint(final_mlr))
mlr_ci$Variable <- rownames(mlr_ci)

mlr_table <- cbind(mlr_table, mlr_ci[,1:2])

colnames(mlr_table)[ncol(mlr_table)-1] <- "CI_Lower"
colnames(mlr_table)[ncol(mlr_table)] <- "CI_Upper"

```


```{r}
#| include: false

# TABLE FOR MLR STATS

## GETTING STATS

mlr_stats <- data.frame(
  Statistic = c("R-squared", "Adjusted R-squared", "N"),
  Value = c(
    round(mlr_summary$r.squared, 3),
    round(mlr_summary$adj.r.squared, 3),
    nobs(final_mlr)
  )
)

```


The model explains approximately 14.1% of the variability in log(BMI) (R² = 0.141), indicating a modest fit (@tbl-mlr-stat). This suggests that a substantial portion of variability in BMI remains unexplained by the predictors included in the model. Overall, physical activity and general health emerge as the most important factors associated with BMI in this model, while smoking and age show weaker or more limited associations after adjustment of other confounders.

```{r}
#| echo: false
#| label: tbl-mlr-stat
#| tbl-cap: "Multiple Linear Regression Statistics"

mlr_stats %>% 
  kable(col.names = c("Statistic", "Value"))

```


```{r}
#| include: false

# MLR STATISTICS
  # observations = 163198.000
  # r2 & adjr2 = 0.145
```


{{< pagebreak >}}

### Predicting Heart Attack Risk

#### Descriptive Statistics & Model Results

Individuals with and without heart disease exhibit distinct differences in the proportions of race, sex, age, health behaviors (smoking, alcohol consumption), and medical conditions (diabetes, kidney disease, asthma, skin cancer, stroke, and difficulty walking). Of individuals with heart disease, 59.0% were female, 82.2% were White, and 87.6% were 55 years or older. These proportions decrease for individuals without heart disease, of whom only 46.5% were female, 76.2% were White, and 51.4% were 55 years or older (@fig-heartdisease-race; @fig-heartdisease-sex; @fig-heartdisease-age).

The majority (58.6%) of individuals with heart disease were smokers, while smokers constituted the minority (39.6%) of individuals without heart disease (@fig-heartdisease-smoking). Interestingly, a slightly larger portion (7.1%) of those without heart disease reported drinking alcohol compared to those with heart disease (4.2%) (@fig-heartdisease-alcohol).

In general, a greater percentage (14.1%) of those with heart disease were categorized as having 'poor' health compared to those without heart disease (2.5%) (@fig-heartdisease-gen-health). More specifically, comorbidity with other health complications was more common amongst those with heart disease. Of individuals with heart disease, 33.1% were diabetic, 12.6% had kidney disease, 18.0% has asthma, 18.2% had skin cancer, 16.0% had had a stroke, and 36.6% experienced difficulty walking. These percentages were all lower for individuals with out heart disease: Only 11.7% were diabetic, 2.9% had kidney disease, 13.0% had asthma, 8.5% had skin cancer, 2.6% had had a stroke, and 11.8% experienced difficulty walking (@fig-heartdisease-diabetes; @fig-heartdisease-kidney; @fig-heartdisease-cancer; @fig-heartdisease-asthma; @fig-heartdisease-stroke).

```{r}
#| include: false

# calculating proportions

# sex
heartdisease |>
  count(heart_disease, sex) |>
  group_by(heart_disease) |>
  mutate(prop = n / sum(n))

# race
heartdisease |>
  count(heart_disease, race) |>
  group_by(heart_disease) |>
  mutate(prop = n / sum(n))

# age
heartdisease |>
  count(heart_disease, age_category) |>
  group_by(heart_disease) |>
  mutate(prop = n / sum(n))


# smoking
heartdisease |>
  count(heart_disease, smoking) |>
  group_by(heart_disease) |>
  mutate(prop = n / sum(n))

# alcohol
heartdisease |>
  count(heart_disease, alcohol_drinking) |>
  group_by(heart_disease) |>
  mutate(prop = n / sum(n))

# general health
heartdisease |>
  count(heart_disease, gen_health) |>
  group_by(heart_disease) |>
  mutate(prop = n / sum(n))

# diabetes
heartdisease |>
  count(heart_disease, diabetic) |>
  group_by(heart_disease) |>
  mutate(prop = n / sum(n))

# kidney disease
heartdisease |>
  count(heart_disease, kidney_disease) |>
  group_by(heart_disease) |>
  mutate(prop = n / sum(n))

# asthma
heartdisease |>
  count(heart_disease, asthma) |>
  group_by(heart_disease) |>
  mutate(prop = n / sum(n))

# skin cancer
heartdisease |>
  count(heart_disease, skin_cancer) |>
  group_by(heart_disease) |>
  mutate(prop = n / sum(n))

# stroke
heartdisease |>
  count(heart_disease, stroke) |>
  group_by(heart_disease) |>
  mutate(prop = n / sum(n))

# difficulty walking
heartdisease |>
  count(heart_disease, diff_walking) |>
  group_by(heart_disease) |>
  mutate(prop = n / sum(n))
```



The significance of these proportion differences is validated in the outcome of our logistic regression. The ultimate results of our logistic regression on the odds of heart disease are captured in @fig-log-reg-coefficients (See @tbl-log-reg-coeffs in the appendix for details on all coefficient estimates). All predictors included in our final model significantly (p < 0.05) impacted the odds of an individual experiencing heart disease, excluding the following: mental health score, being Native American, and being an unspecified race.


```{r}
#| echo: false
#| fig-cap: "Coefficients of logistic regression model. Blue data points represent the additive effect of each predictor on the log-odds of heart disease. Red dashed line represents no effect on the log-odds of heart disease. Bars represent 95% confidence interval for the estimate. All effects are significant (p < 0.05) aside from the terms Mental Health, Native American, and Other Race."
#| label: fig-log-reg-coefficients
log_reg_tidy <- tidy(final_log_reg, conf.int = TRUE) |>
  filter(term != "(Intercept)") |>
  mutate(
    term = case_match(term,
                     "alcohol_drinking1" ~ "Drinks Alcohol",
                     "mental_health" ~ "Mental Health",
                     "sex1" ~ "Male",
                     "age_category3" ~ "55 and older",
                     "raceAsian" ~ "Asian",
                     "raceHispanic" ~ "Hispanic",
                     "diabetic1" ~ "Diabetic",
                     "gen_healthGood" ~ "Good Health",
                     "asthma1" ~ "Asthma",
                     "skin_cancer1" ~ "Skin Cancer",
                     "smoking1" ~ "Smoker",
                     "stroke1" ~ "Stroke",
                     "diff_walking1" ~ "Difficulty Walking",
                     "age_category2" ~ "35-54 years old",
                     "raceAmerican Indian/Alaskan Native" ~ "Native American",
                     "raceBlack" ~ "Black",
                     "raceOther" ~ "Other Race",
                     "gen_healthFair" ~ "Fair Health",
                     "gen_healthExcellent" ~ "Excellent Health",
                     "kidney_disease1" ~ "Kidney Disease",
                     .default = term
                     ),
    term = reorder(term, estimate)
  )

ggplot(log_reg_tidy, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high), width = 0.2) +
  geom_point(size = 2, color = "steelblue") +
  labs(
    x = "Effect on log-odds of heart disease (with 95% confidence interval)",
    y = NULL
  ) + 
  theme_classic()

```




```{r}
#| include: false

# create modified data set with categorical sleep, age, & bmi variables

young = c("18-24", "25-29", "30-34", "35-39")
middle_age = c("40-44", "45-49", "50-54", "55-59")
old = c("60-64", "65-69", "70-74", "75-79", "80 or older")

disease2 <- disease_data |>
  mutate(sleep_time = case_when(
    sleep_time <= 10 & sleep_time >= 7 ~ "Healthy Sleep (7-10 hrs)",
    sleep_time < 7 ~ "Sleep Deprived (<7 hrs)",
    sleep_time > 10 ~ "Extra Sleep (>10 hrs)",
    TRUE ~ NA_character_),
    
    bmi = case_when(
    bmi < 19 ~ "Underweight",
    bmi >= 19 & bmi < 25 ~ "Normal Weight",
    bmi >= 25 & bmi < 30 ~ "Overweight",
    bmi >= 30 ~ "Obese",
    TRUE ~ NA_character_),
    
   # age_category = case_when(
 #   age_category %in% young ~ "Young (18-39)",
  #  age_category %in% middle_age ~ "Middle Aged (40-59)",
  #  age_category %in% old ~ "Elderly (60-80+)",
  #  TRUE ~ NA_character_)
    
    )
```

```{r}
#| include: false

# testing for correlation

#cor(disease_data$smoking, disease_data$asthma) #0.02414853
#cor(disease_data$diabetic, disease_data$heart_disease) #0.1747823

```

[Demographics]{.underline}

Compared to White peers, and holding all other variables constant, our model predicts that Hispanic, Black, and Asian individuals have 0.75 (95% CI: \[0.70, 0.81\]), 0.70 (95% CI: \[0.66, 0.76\]), and 0.60 (95% CI: \[0.51, 0.71\]) times the odds of having heart disease, respectively (@fig-heartdisease-race; @tbl-log-reg-coeffs). In short, White individuals are predicted to have the highest likelihood of heart attack, contrary to our initial hypothesis which posited that historical discrimination against non-White communities in the US would lead to worse health outcomes (i.e. greater risk for heart disease).

As one may expect, the predicted risk for heart disease increases with age. Compared to 18-34 year olds of the same sex, race, and health conditions, 35-54 year olds are expected to have 2.97 the odds of heart disease (95% CI: [2.61, 3.37]). This risk increases even more dramatically for ages 55 years and older, whose odds of heart disease are predicted to be 8.9 times (95% CI: [7.89, 10.06]) that of 18-34 year olds (@fig-heartdisease-age; @tbl-log-reg-coeffs).

Sex also has a significant impact on the odds of having heart disease. Males have 0.55 times (95% CI: [0.53, 0.57]) the odds of have heart disease compared to females, holding all other predictors constant (@fig-heartdisease-sex; @tbl-log-reg-coeffs).

:::::: {layout="[0.3, 0.3, 0.3]"}
::: {#plot1}
```{r}
#| echo: false
#| label: fig-heartdisease-race
#| fig-cap: "Race"
## DEMOGRAPHICS

# RACE
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = race)) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Race") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(values = c(
    "White" = "antiquewhite",
    "American Indian/Alaskan Native" = "tomato",
    "Asian" = "skyblue",
    "Black" = "orange",
    "Hispanic" = "darkolivegreen3",
    "Other" = "plum"),
    labels = c(
      "American Indian/Alaskan Native" = "Native American"
      )
    ) +
  theme_classic() + 
  theme(legend.position = "bottom")
```
:::

::: {#plot2}
```{r}
#| echo: false
#| label: fig-heartdisease-sex
#| fig-cap: "Sex"
# SEX
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = factor(sex))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Sex") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    labels = c("1" = "Male", "0" = "Female"),
    values = c("1" = "tomato", "0" = "skyblue")) +
  theme_classic() + 
  theme(legend.position = "bottom")
```
:::

::: {#plot3}
```{r}
#| echo: false
#| label: fig-heartdisease-age
#| fig-cap: "Age"

# AGE
ggplot(data = disease2, aes(x = factor(heart_disease), fill = age_category)) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease",
       y = "Count",
       fill = "Age Group") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("1" = "orange", "2" = "tomato", "3" = "skyblue"),
    labels = c("18-34", "35-54", "55 and older")) +
  theme_classic() + 
  theme(legend.position = "bottom")




```
:::
:::::

```{r}
#| include: false
# AGE
ggplot(data = disease_data, aes(x = age_category, fill = factor(heart_disease))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Age",
       y = "Count",
       fill = "Heart Disease Status") +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  scale_x_discrete(labels = c("1" = "18-34",
                              "2" = "35-54",
                              "3" = "55 and older")) +
  theme_classic() + 
  theme(legend.position = "bottom")
```


[Behavior]{.underline}


As we expected, our model predicts that individuals who smoke are at higher risk for heart disease. In particular, their odds of having a heart attack are 1.41 times (95% CI: [1.35, 1.46]) the odds for individuals who do not smoke, holding all other predictors constant (@fig-heartdisease-smoking; @tbl-log-reg-coeffs). Interestingly, alcohol consumption is predicted to have a significant negative effect on the risk of heart disease. Individuals who drink alcohol have 0.70 times (95% CI: [0.64, 0.76]) the odds of heart disease compared to individuals who do not drink alcohol (@fig-heartdisease-alcohol; @tbl-log-reg-coeffs).

:::: {layout="[0.45, 0.45]"}
:::{#plot1}
```{r}
#| echo: false
#| label: fig-heartdisease-smoking
#| fig-cap: "Smoking"

# SMOKING
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = factor(smoking))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Smoker") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() + 
  theme(legend.position = "bottom")



```
:::
:::{#plot2}
```{r}
#| echo: false
#| label: fig-heartdisease-alcohol
#| fig-cap: "Alcohol Consumption"

# ALCOHOL
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = factor(alcohol_drinking))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Drinks Alcohol") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() + 
  theme(legend.position = "bottom")
```
:::
::::


[Medical Conditions]{.underline}

Overall, individuals categorized as having 'Fair', 'Good', and 'Excellent' general health were predicted to have a lower risk of heart disease. Compared to the odds of heart disease for those with 'Poor' general health, individuals in the other health categories had 0.68 (95% CI: [0.64, 0.72]), 0.42 (95% CI: [0.40, 0.45]), and 0.13 (95% CI: [0.12, 0.14]) times the odds of heart attack, respectively (@fig-heartdisease-gen-health; @tbl-log-reg-coeffs).



```{r}
#| echo: false
#| label: fig-heartdisease-gen-health
#| fig-width: 4
#| fig-height: 3

# GENERAL HEALTH
heartdisease |>
  filter(!is.na(gen_health)) |>
  ggplot(aes(x = factor(heart_disease), fill = factor(gen_health))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "General Health") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(values = c(
    "Poor" = "tomato",
    "Fair" = "orange",
    "Good" = "gold",
    "Excellent" = "#228B22")) +
  theme_classic() + 
  theme(legend.position = "right")
```

On the other hand, pre-existing health complications such as diabetes, kidney disease, skin cancer, asthma, stroke, and difficulty walking significantly increased the odds of heart disease. Diabetic patients have 1.61 times (95% CI: [1.55, 1.68]) the odds, patients with skin cancer have 1.33 times (95% CI: [1.27, 1.40]) the odds, patients with kidney disease have 1.87 times (95% CI: [1.76, 1.98]) the odds, patients with asthma have 1.25 times (95% CI: [1.19, 1.31]) the odds,  patients who have had a stroke have 2.96 times (95% CI: [2.81, 3.13]) the odds, and patients with difficulty walking have 1.32 (95% CI: [1.27, 1.38]) the odds of heart attack compared to individuals without those respective health conditions, holding all other predictors constant (@fig-heartdisease-diabetes; @fig-heartdisease-kidney; @fig-heartdisease-cancer; @fig-heartdisease-asthma; @fig-heartdisease-stroke @tbl-log-reg-coeffs). Overall, having a stroke increases an individuals risk for heart attack to the highest degree of the health conditions mentioned. 

:::: {layout="[0.3, 0.3, 0.3]"}
:::{#plot1}
```{r}
#| echo: false
#| label: fig-heartdisease-diabetes
#| fig-cap: "Diabetes"

## MEDICAL CONDITIONS

# DIABETES
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = factor(diabetic))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Diabetic") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() + 
  theme(legend.position = "bottom")
```
:::


:::{#plot2}
```{r}
#| echo: false
#| label: fig-heartdisease-kidney
#| fig-cap: "Kidney Disease"

# KIDNEY DISEASE
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = factor(kidney_disease))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Kidney Disease") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() + 
  theme(legend.position = "bottom")
```
:::

:::{#plot3}

```{r}
#| echo: false
#| label: fig-heartdisease-asthma
#| fig-cap: "Asthma"
# ASTHMA
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = factor(asthma))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Asthma") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() + 
  theme(legend.position = "bottom")
```
:::
::::


:::: {layout="[0.3, 0.3, 0.3]"}

:::{#plot2}

```{r}
#| echo: false
#| label: fig-heartdisease-cancer
#| fig-cap: "Skin Cancer"
# SKIN CANCER
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = factor(skin_cancer))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Skin Cancer") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() + 
  theme(legend.position = "bottom")

```
:::

:::{#plot3}
```{r}
#| echo: false
#| label: fig-heartdisease-stroke
#| fig-cap: "Stroke"
# STROKE
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = factor(stroke))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Stroke") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() + 
  theme(legend.position = "bottom")
```
:::

:::{#plot1}
```{r}
#| echo: false
#| label: fig-heartdisease-walking
#| fig-cap: "Difficulty Walking"
# DIFFICULTY WALKING
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = factor(diff_walking))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Difficulty Walking") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() + 
  theme(legend.position = "bottom")
```

:::

::::





#### Regression Analysis

```{r}
#| include: false

# calculate the odds & probability
heart_test <- heart_test |>
  mutate(
    log_odds = predict(final_log_reg, newdata = heart_test, type = "link"),
    odds = exp(log_odds),
    prob = odds / (1 + odds)
  )

```


To assess the sensitivity and specificity of our logistic regression model and to determine an appropriate probability threshold for classification, we applied our model to a test data set (see Methodology for how we created this split) and plotted a Receive Operating Characteristic (ROC) curve (@fig-roc). The Area Under the Curve (AUC) is 0.83, indicating that the model's classification of individuals with and without heart disease is much better than random (AUC = 0.5, represented by the y = x dashed line). We opted to use Youden's J statistic  to determine a probability threshold. This statistic identifies the maximum vertical distance between the ROC curve and the $y = x$ line, thereby maximizing the difference between the true positive rate (sensitivity) and the false positive rate (1 - specificity). We obtained a Youden J statistic of 0.09886 and set this as our threshold for subsequent analysis.
```{r}
#| echo: false
#| label: fig-roc
#| fig-cap: "Model Performance: Receiver Operating Characteristic (ROC) Curve for predicting heart disease. The dotted red line represents a model that is no better than random chance. Performance was assessed by fitting the model on a test data set."
#| fig-width: 3
#| fig-height: 2.5

# calculate predicted probabilities
prob <- predict(final_log_reg, newdata = heart_test, type = "response")

roc <- roc(heart_test$heart_disease, prob)
df <- data.frame(sensitivity = roc$sensitivities, specificity = roc$specificities)

ggplot(data = df, aes(x = 1 - specificity, y = sensitivity)) +
  geom_smooth(color = "black", linewidth = 0.6) +
  labs(
    x = "1 - Specificity (False Positive Rate)",
    y = "Sensitivity (True Positive Rate)") + 
  geom_abline(slope = 1,
              intercept = 0, 
              linetype = "dotted", 
              color = "red") +
  theme_classic()


# calculate the Area Under the Curve
#auc(roc) # 0.8292

# pick a threshold
# Method 1: Youden's J statistic (maximum vertical distance between ROC curve & diagonal line)
  # maximizes the diff between true positive rate & false positive rate
youden <- coords(roc, "best", best.method = "youden")
  # threshold = 0.09886

# Method 2: threshold closest to (0, 1) (perfect classification)
topleft <- coords(roc, "best", best.method = "closest.topleft")
  # threshold = 0.1161

```

```{r}
#| include: false
## another method for ROC curve visualization

# get predictions from original data
log_preds <- augment(final_log_reg, type.predict = "response") |>
  mutate(heart_disease = as.factor(heart_disease))

# calculate ROC curve coordinates
roc_data <- roc_curve(log_preds, heart_disease, .fitted, event_level = "second")

# plot
autoplot(roc_data) +
  labs(
    title = "Model Performance: ROC Curve",
    subtitle = "Predicting Heart Disease",
    caption = "Diagonal line represents a model which is no better than random chance."
  ) + 
  theme_classic()
```

Using this threshold, we constructed a confusion matrix for the predicted heart disease outcomes on our test data (see section 2 for details on how we developed the train and test data) (@tbl-confusion-matrix). From this confusion matrix, we then calculated the rates for true positive, false positive, true negative, and false negative detection (@tbl-confusion-matrix-rates). At the threshold of 0.09886, our model can accurately identify 80.6% of individuals with heart disease, indicating high model sensitivity. The model's ability to accurately classify individuals as not having heart disease is also quite high at 70.2%, indicating high model specificity. This being said, the model will still incorrectly diagnose an individual has having heart disease 29.8% of the time and fails to properly identify a person as having heart disease 19.4% of the time.

::::: {layout="[0.5, 0.3]"}
:::{#plot1}
```{r}
#| echo: false
#| label: tbl-confusion-matrix
#| tbl-cap: "Confusion Matrix for Test Data"


# classify heart disease based on probability
threshold <- youden$threshold
pred_class <- ifelse(prob > threshold, 1, 0)

# create confusion matrix
confusion_matrix <- table(Predicted = pred_class, Actual = heart_test$heart_disease) 

rownames(confusion_matrix) <- c("No Heart Disease (Pred)", "Heart Disease (Pred)")
colnames(confusion_matrix) <- c("No Heart Disease (Actual)", "Heart Disease (Actual)")

confusion_matrix|> kable()
```
:::

:::{#plot2}
```{r}
#| echo: false
#| label: tbl-confusion-matrix-rates

# calculate true positive, false positive, false negative, and true negative rates
TN <- confusion_matrix[1, 1]
FN <- confusion_matrix[1, 2]
FP <- confusion_matrix[2, 1]
TP <- confusion_matrix[2, 2]

# measures of classification error
FNR <- FN/(TP + FN)
FPR <- FP/(TN + FP)

# measures of classification ability
TPR <- TP/(TP + FN) # sensitivity
TNR <- TN/(TN + FP) # specificity

# make a table
rates_tbl <- data.frame(
  metric = c("True Positive Rate", 
             "False Positive Rate",
             "False Negative Rate",
             "True Negative Rate"),
  value = c(TPR, FPR, FNR, TNR)
)

rates_tbl |> kable()

```
:::
:::::

```{=html}
<!-- 
If you have more than 2 research questions, try to group them into no 
more than 2 groups. Rename the subsections to give a short description of the
group of questions (e.g. "Landscape of US colleges" and "Factors associated with dropout")
-->
```
