```{r}
#| include: false

# loading libraries
library(tidyverse)
library(tidyr)
library(kableExtra)
library(pROC)

# load data
disease_data <- heartdisease

```

## Results

### Factors Associated with BMI ....

#### Descriptive Statistics

**Demographics**

insert text

**Behaviors**

insert texttt

**Health**

insert texttt

#### Regression Analysis

### Predicting Heart Attack Risk ....

#### Descriptive Statistics

```{r}
#| include: false

# create modified data set with categorical sleep, age, & bmi variables

young = c("18-24", "25-29", "30-34", "35-39")
middle_age = c("40-44", "45-49", "50-54", "55-59")
old = c("60-64", "65-69", "70-74", "75-79", "80 or older")

disease2 <- disease_data |>
  mutate(sleep_time = case_when(
    sleep_time <= 10 & sleep_time >= 7 ~ "Healthy Sleep (7-10 hrs)",
    sleep_time < 7 ~ "Sleep Deprived (<7 hrs)",
    sleep_time > 10 ~ "Extra Sleep (>10 hrs)",
    TRUE ~ NA_character_),
    
    bmi = case_when(
    bmi < 19 ~ "Underweight",
    bmi >= 19 & bmi < 25 ~ "Normal Weight",
    bmi >= 25 & bmi < 30 ~ "Overweight",
    bmi >= 30 ~ "Obese",
    TRUE ~ NA_character_),
    
   # age_category = case_when(
 #   age_category %in% young ~ "Young (18-39)",
  #  age_category %in% middle_age ~ "Middle Aged (40-59)",
  #  age_category %in% old ~ "Elderly (60-80+)",
  #  TRUE ~ NA_character_)
    
    )
```

```{r}
#| include: false

# testing for correlation

#cor(disease_data$smoking, disease_data$asthma) #0.02414853
#cor(disease_data$diabetic, disease_data$heart_disease) #0.1747823

```

[Demographics]{.underline}

Compared to White peers, and holding all other variables constant, our model predicts with that Hispanic, Black, and Asian individuals have 0.75 (95% CI: \[0.70, 0.81\]), 0.70 (95% CI: \[0.66, 0.76\]), and 0.60 (95% CI: \[0.51, 0.71\]) times the odds of having heart disease, respectively (@fig-heartdisease-race; @tbl-log-reg-coeffs) . In short, White individuals are predicted to have the highest likelihood of heart attack.

As one may expect, the predicted risk for heart disease increases with age. Compared to 18-34 year olds of the same sex, race, and health conditions, 35-54 year olds are expected to have 2.97 the odds of heart disease (95% CI: [2.61, 3.37]). This risk increases even more dramatically for ages 55 years and older, whose odds of heart diseae are predicted to be 8.9 times (95% CI: [7.89, 10.06]) that of 18-34 year olds (@fig-heartdisease-age; @tbl-log-reg-coeffs).

Sex also has a significant impact on the odds of having heart disease. **SEX 1** has 0.55 times (95% CI: [0.53, 0.57]) the odds of have heart disease compared to **SEX 0**, holding all other predictors constant (@fig-heartdisease-sex; @tbl-log-reg-coeffs).

:::::: {layout="[0.3, 0.3, 0.3]"}
::: {#plot1}
```{r}
#| echo: false
#| label: fig-heartdisease-race
## DEMOGRAPHICS

# RACE
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = race)) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Race") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  #scale_fill_manual(values = c(1 = "tomato", 0 = "skyblue")) +
  theme_classic() + 
  theme(legend.position = "bottom")
```
:::

::: {#plot2}
```{r}
#| echo: false
#| label: fig-heartdisease-sex
# SEX
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = factor(sex))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Sex") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  #scale_fill_manual(values = c(1 = "tomato", 0 = "skyblue")) +
  theme_classic() + 
  theme(legend.position = "bottom")
```
:::

::: {#plot3}
```{r}
#| echo: false
#| label: fig-heartdisease-age

# AGE
ggplot(data = disease2, aes(x = factor(heart_disease), fill = age_category)) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease",
       y = "Count",
       fill = "Age Group") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("1" = "orange", "2" = "tomato", "3" = "skyblue"),
    labels = c("18-34", "35-54", "55 and older")) +
  theme_classic() + 
  theme(legend.position = "bottom")




```
:::
::::::
```{r}
#| include: false
# AGE
ggplot(data = disease_data, aes(x = age_category, fill = factor(heart_disease))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Age",
       y = "Count",
       fill = "Heart Disease Status") +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  scale_x_discrete(labels = c("1" = "18-34",
                              "2" = "35-54",
                              "3" = "55 and older")) +
  theme_classic() + 
  theme(legend.position = "bottom")
```

:::::: {layout="[0.3, 0.3, 0.3]"}
<div>

</div>

<div>

</div>

<div>

</div>
::::::


</div>

<div>


[Behavior]{.underline}


As we expected, our model predicts that individuals who smoke are at higher risk for heart disease. In particular, their odds of having a heart attack are 1.41 times (95% CI: [1.35, 1.46]) the odds for individuals who do not smoke, holding all other predictors constant (@fig-heartdisease-smoking; @tbl-log-reg-coeffs). Interestingly, alcohol consumption is predicted to have a significant negative effect on the risk of heart disease. Individuals who drink alcohol have 0.70 times (95% CI: [0.64, 0.76]) the odds of heart disease compared to individuals who do not drink alcohol (@fig-heartdisease-alcohol; @tbl-log-reg-coeffs).

```{r}
#| echo: false
#| label: fig-heartdisease-smoking

# SMOKING
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = factor(smoking))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Smoker") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() + 
  theme(legend.position = "bottom")



```

```{r}
#| echo: false
#| label: fig-heartdisease-alcohol

# ALCOHOL
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = factor(alcohol_drinking))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Drinks Alcohol") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() + 
  theme(legend.position = "bottom")
```

</div>
::::::


[Medical Conditions]{.underline}


:::::: {layout="[0.3, 0.3, 0.3]"}
<div>

```{r}
#| echo: false

## MEDICAL CONDITIONS

# DIABETES
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = factor(diabetic))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Diabetic") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() + 
  theme(legend.position = "bottom")
```

</div>

<div>

```{r}
#| echo: false

# KIDNEY DISEASE
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = factor(kidney_disease))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Kidney Disease") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() + 
  theme(legend.position = "bottom")
```

</div>

<div>

```{r}
#| echo: false
# ASTHMA
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = factor(asthma))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Asthma") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() + 
  theme(legend.position = "bottom")
```

</div>
::::::

:::::: {layout="[0.3, 0.3, 0.3]"}
<div>

```{r}
#| echo: false
# SKIN CANCER
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = factor(skin_cancer))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Skin Cancer") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() + 
  theme(legend.position = "bottom")

```

</div>

<div>

```{r}
#| echo: false
# STROKE
ggplot(data = disease_data, aes(x = factor(heart_disease), fill = factor(stroke))) +
  geom_bar(color = "black", position = "fill") + 
  labs(x = "Heart Disease Status",
       y = "Count",
       fill = "Stroke") +
  scale_x_discrete(labels = c("0" = "No Heart Disease",
                              "1" = "Heart Disease")) +
  scale_fill_manual(
    values = c("0" = "skyblue", "1" = "tomato"),
    labels = c("No", "Yes")
  ) +
  theme_classic() + 
  theme(legend.position = "bottom")
```

</div>

<div>

</div>
::::::

#### Regression Analysis

-   final regression equation

```{r}
#| echo: false
#| label: tbl-log-reg-coeffs
#| tbl-cap: "Logistic Regression Coefficients"
reg_sum <- summary(final_log_reg)

coefficients <- as.data.frame(reg_sum$coefficients)
coeff_sorted <- coefficients[order(coefficients$Estimate, decreasing = TRUE), ]


# calculate the odds & probability
coeff_sorted <- coeff_sorted |>
  mutate(
    odds_coeff = exp(coeff_sorted$Estimate),
    prob = odds_coeff / (1 + odds_coeff),
    ci_low = exp(Estimate - 2*`Std. Error`),
    ci_high = exp(Estimate + 2*`Std. Error`)
  )

colnames(coeff_sorted) <- c("Estimate (log-odds)", "Std. Error", "Z value", "P-value", "Estimate (odds)", "Probability", "CI low (odds)", "CI high (odds)")

coeff_sorted |> kable() |>
  kable_styling() |>
  row_spec(which(coeff_sorted$`Estimate (log-odds)` >= 0),  background = "#ffe6e6") |>
  row_spec(which(coeff_sorted$`Estimate (log-odds)` < 0), background = "#e6ffe6") |>
  row_spec(which(coeff_sorted$`P-value` > 0.05), background = "#c7d1d9")
```

-   interpretation of coefficients

-   odds interpretation (odds ratio, confidence interval)

-   probability interpretation

```{r}
#| include: false


# calculate the odds & probability
heart_test <- heart_test |>
  mutate(
    log_odds = predict(final_log_reg, newdata = heart_test, type = "link"),
    odds = exp(log_odds),
    prob = odds / (1 + odds)
  )

```

```{r}
#| echo: false
#| label: fig-roc
#| fig-cap: "Receiver Operating Characteristic (ROC) Curve"

# calculate predicted probabilities
prob <- predict(final_log_reg, newdata = heart_test, type = "response")

roc <- roc(heart_test$heart_disease, prob)
df <- data.frame(sensitivity = roc$sensitivities, specificity = roc$specificities)

ggplot(data = df, aes(x = 1 - specificity, y = sensitivity)) +
  geom_point() +
  labs(
    x = "1 - Specificity (False Positive Rate)",
    y = "Sensitivity (True Positive Rate)") + 
  geom_abline(slope = 1,
              intercept = 0, 
              linetype = "dashed", 
              color = "steelblue") +
  theme_classic()


# calculate the Area Under the Curve
auc(roc) # 0.8292

# pick a threshold
# Method 1: Youden's J statistic (maximum vertical distance between ROC curve & diagonal line)
  # maximises the diff between true positive rate & false positive rate
youden <- coords(roc, "best", best.method = "youden")
  # threshold = 0.09886

# Method 2: threshold closest to (0, 1) (perfect classification)
topleft <- coords(roc, "best", best.method = "closest.topleft")
  # threshold = 0.1161

```

::::: {layout="[0.5, 0.4]"}
<div>

```{r}
#| echo: false
#| tbl-cap: "Confusion Matrix for Predicted Heart Disease Outcomes on Test Data"


# classify heart disease based on probability
threshold <- youden$threshold
pred_class <- ifelse(prob > threshold, 1, 0)

# create confusion matrix
confusion_matrix <- table(Predicted = pred_class, Actual = heart_test$heart_disease) 

rownames(confusion_matrix) <- c("No Heart Disease (Pred)", "Heart Disease (Pred)")
colnames(confusion_matrix) <- c("No Heart Disease (Actual)", "Heart Disease (Actual)")

confusion_matrix|> kable()
```

</div>

<div>

```{r}
#| echo: false

# calculate true positive, false positive, false negative, and true negative rates
TN <- confusion_matrix[1, 1]
FN <- confusion_matrix[1, 2]
FP <- confusion_matrix[2, 1]
TP <- confusion_matrix[2, 2]

# measures of classification error
FNR <- FN/(TP + FN)
FPR <- FP/(TN + FP)

# measures of classification ability
TPR <- TP/(TP + FN) # sensitivity
TNR <- TN/(TN + FP) # specificity

# make a table
rates_tbl <- data.frame(
  metric = c("True Positive Rate", 
             "False Positive Rate",
             "False Negative Rate",
             "True Negative Rate"),
  value = c(TPR, FPR, FNR, TNR)
)

rates_tbl |> kable()

```

</div>
:::::

```{=html}
<!-- 
If you have more than 2 research questions, try to group them into no 
more than 2 groups. Rename the subsections to give a short description of the
group of questions (e.g. "Landscape of US colleges" and "Factors associated with dropout")
-->
```
