---
title: "Advanced Quarto Layouts"
subtitle: "Columns, Callouts, and Tabsets"
author: "Data Science Student"
format: 
  html:
    theme: cosmo
    toc: true
    code-fold: true
execute:
  message: false
  warning: false    
---
```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(broom)
library(scales)
```

# 1. Using Columns for Data Summaries
Columns are perfect for placing descriptive text alongside a quick data snapshot.

::: {.columns}

::: {.column width="45%"}
### Data Overview
The `mtcars` dataset contains engine and performance specs for 32 automobiles. 

**Key Variables:**
- `mpg`: Miles per gallon
- `wt`: Weight (1000 lbs)
- `am`: Transmission (0 = auto, 1 = manual)
:::

::: {.column width="10%"}
:::

::: {.column width="45%"}
### Quick Glimpse
```{r}
mtcars |> 
  select(mpg, wt, am) |> 
  head(5) |> 
  knitr::kable()
```
:::

:::

---

# 2. Using Callouts for Critical Insights
Callouts draw the reader's attention to specific types of information.

::: {.callout-note}
## Research Objective
We aim to model fuel efficiency while accounting for vehicle weight and transmission type.
:::

::: {.callout-warning}
## Data Limitation
This dataset is from 1974. Technological advances in modern engines may result in different coefficient scales today.
:::

::: {.callout-tip icon=false}
## 💡 Pro-Tip
When plotting probabilities, always use `scales::label_percent()` for the Y-axis to make it more readable for non-statisticians.
:::

---

# 3. Using Tabsets for Model Comparison
Tabsets are the best way to present multiple models without overwhelming the reader.

::: {.panel-tabset}

## 🚗 Linear Model (MPG)
This model predicts Miles Per Gallon.

```{r}
#| label: fig-mlr-report
#| fig-cap: "Linear coefficients showing that Weight is the strongest predictor of MPG."

m_lin <- lm(mpg ~ wt + hp + am, data = mtcars)

tidy(m_lin, conf.int = TRUE) |> 
  filter(term != "(Intercept)") |> 
  mutate(term = reorder(term, estimate)) |> 
  ggplot(aes(x = estimate, y = term)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high), width = 0.2) +
  geom_point(size = 3, color = "steelblue") +
  theme_minimal()
```

## ⚙️ Logistic Model (Transmission)
This model predicts the probability of a car being a Manual.

```{r}
#| label: fig-log-report
#| fig-cap: "Probability of manual transmission decreases sharply as weight increases."

m_log <- glm(am ~ wt, data = mtcars, family = binomial)

# Generate smooth prediction line
grid <- data.frame(wt = seq(min(mtcars$wt), max(mtcars$wt), length.out = 100))
grid$prob <- predict(m_log, newdata = grid, type = "response")

ggplot(grid, aes(x = wt, y = prob)) +
  geom_line(color = "darkgreen", linewidth = 1) +
  scale_y_continuous(labels = label_percent()) +
  labs(x = "Weight (1000 lbs)", y = "Prob(Manual)") +
  theme_minimal()
```

:::

---

# Summary Checklist
1. **Columns:** Use `::: {.columns}` and `::: {.column width="X%"}` to place items side-by-side.
2. **Callouts:** Use `::: {.callout-TYPE}` (note, warning, tip, important, caution).
3. **Tabsets:** Use `::: {.panel-tabset}`. Each `## Header` inside becomes a new tab.
4. **Closing:** Always remember to close every div with `:::`.