---
title: "Visualizing Regression Results"
subtitle: "Communicating Uncertainty with ggplot2"
author: "Template"
format: 
  html:
    toc: true
theme: cosmo
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(broom)    # For tidying model objects
library(modelr)   # For generating prediction grids
library(scales)   # For better axis labels
library(yardstick) # For ROC curve
```

## Introduction

When presenting regression results, tables are informative but visualizations are often more intuitive for understanding the magnitude of effects and the precision of our estimates (Confidence Intervals).

## 1. The Forest Plot: Visualizing Coefficients

The "Forest Plot" shows the point estimate of each coefficient and its 95% confidence interval. This is much better than looking at p-values alone.

### Step A: Fit the Model 🔧

We will use the built-in `mtcars` dataset to model Miles Per Gallon (MPG).

```{r}
# Fit a multiple linear regression
m_lin <- lm(mpg ~ hp + wt + am + cyl, data = mtcars)

```

### Step B: Prepare the Results 👔

We start by tidying the results and including confidence intervals

```{r}
results_tidy <- tidy(m_lin, conf.int = TRUE) 
head(results_tidy)
```

In this dataset, each variable in the model is a row: `term` is the variable's name, `estimate` is its estimated coefficient, `conf.low` and `conf.high` are the lower and upper bounds of its confidence interval.

We want to plot the confidence intervals of each variable so we prepare the dataset to look nice in the plot:

```{r}
results_tidy <- results_tidy |> 
   # Usually we exclude the intercept for plotting
  filter(term != "(Intercept)") |> 
  # We give "nice" names for the plot axis
  mutate(
    term = case_match(term,
      "hp"  ~ "Horsepower",
      "wt"  ~ "Weight (1000 lbs)",
      "am"  ~ "Transmission (Manual)",
      "cyl" ~ "Cylinders",
      .default = term
    ),
  # We turn 'term' into a factor sorted by the 'estimate' value.
  # This way, the coefficients will be shown in order from smallest to largest estimate
    # Without this, ggplot would default to alphabetical order.
     term = reorder(term, estimate)
  )
 
```

### Step C: Create the Plot 🧑🏼‍🎨

Next we make the forest plot!

```{r}
#| label: fig-mpg-forest
#| fig-cap: "Impact of Car Characteristics on Fuel Efficiency."

ggplot(results_tidy, aes(x = estimate, y = term)) +
  # Adds a vertical reference line at 0 (null effect)
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  # Use geom_errorbar with xmin and xmax for horizontal intervals
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high), width = 0.2) +
  geom_point(size = 3, color = "steelblue") +
  labs(
    title = "Impact of Car Characteristics on Fuel Efficiency",
    x = "Effect on MPG (with 95% Confidence Interval)",
    y = NULL, # We remove the axis label because variable names are self-explanatory
    caption = "Data source: 1974 Motor Trend US magazine. Model adjusted for engine and weight factors."
  ) +
  theme_minimal()
```

## 2. Visualizing Predictions

While Forest Plots show us *coefficients*, Prediction Plots show us the *real-world relationship* between a predictor and the outcome. We use `modelr::data_grid()` to create a "typical" scenario (e.g., varying weight while holding other variables at their mean).

### MLR

#### Step A: Create the Prediction Data

We want to see how **Weight** affects **MPG**, but we must tell R what to do with the other variables in the model (`hp`, `am`, `cyl`).

Code snippet

```{r}
# 1. Generate a grid of data
# We vary 'wt' across its full range and hold others at their mean/mode
plot_grid <- mtcars |> 
  data_grid(
    wt = seq_range(wt, n = 100),
    hp = mean(hp),
    am = mean(am),
    cyl = mean(cyl)
  )

# 2. Add predictions and standard errors
mlr_preds <- augment(m_lin, newdata = plot_grid, se_fit = TRUE) |> 
  mutate(
    # Calculate 95% Confidence Interval bounds
    lower = .fitted - (1.96 * .se.fit),
    upper = .fitted + (1.96 * .se.fit)
  )
```

#### Step B: Plot the Trend with Uncertainty

A good prediction plot uses a line for the estimate and a "ribbon" for the confidence interval.

```{r}
#| label: fig-mlr-predictions
#| fig-cap: "Predicted MPG based on vehicle weight. The blue line represents the expected MPG, while the shaded area shows the 95% confidence interval."
#| echo: fenced

ggplot(mlr_preds, aes(x = wt, y = .fitted)) +
  # The ribbon represents our uncertainty
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "steelblue", alpha = 0.2) +
  # The line represents our best estimate
  geom_line(color = "steelblue", linewidth = 1) +
  labs(
    title = "The Relationship Between Weight and Fuel Efficiency",
    subtitle = "Adjusted for Horsepower, Transmission, and Cylinders",
    x = "Weight (1000 lbs)",
    y = "Predicted MPG",
    caption = "Confidence ribbon calculated using standard errors from the MLR model."
  ) +
  theme_minimal()
```

### Logistic regression

For Logistic Regression (GLM), it is often hard to interpret "Log-Odds." Plotting the **predicted probability** is much more effective.

#### Step A: Fit Logistic Model

We'll predict whether a car has an automatic (`am=0`) or manual (`am=1`) transmission based on its weight.

```{r}
m_log <- glm(am ~ wt + hp, data = mtcars, family = binomial)
```

#### Step B: Create a Prediction Grid

To plot a smooth line with confidence ribbons, we simulate data across the entire range of our predictor.

```{r}
# Create a grid of weights, holding HP at its mean
grid <- mtcars  |> 
  data_grid(
    wt = seq_range(wt, n = 100),
    hp = mean(hp)
  )

```

We then add estimated log-odds and probabilities. We can use the plogis() to convert log-odds to probabilities, since `plogis()` does

$$p = \frac{1}{1 + e^{-\text{log-odds}}}$$

```{r}
# Add predictions with standard errors
# Note: type = "link" gets us the log-odds scale where SEs are symmetric
preds <- augment(m_log, newdata = grid, se_fit = TRUE) |> 
  mutate(
    # Convert log-odds to probabilities using the inverse logit (plogis)
    prob = plogis(.fitted),
    lower = plogis(.fitted - (1.96 * .se.fit)),
    upper = plogis(.fitted + (1.96 * .se.fit))
  )
```

#### Step C: Plot the Probability Curve

```{r}
#| label: fig-prob-curve
#| fig-cap: "Probability of a manual transmission as a function of car weight, holding horsepower constant at its mean."

ggplot(preds, aes(x = wt, y = prob)) +
  # The Confidence Ribbon
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "grey80", alpha = 0.5) +
  # The Prediction Line
  geom_line(color = "darkgreen", linewidth = 1) +
  # Add the original data points (rug plot)
  geom_rug(data = mtcars, aes(y = am), sides = "b", alpha = 0.5) +
  labs(
    title = "Predicted Probability of Manual Transmission",
    subtitle = "Holding Horsepower at its mean",
    x = "Weight (1000 lbs)",
    y = "Probability (Manual)",
    caption = "The grey ribbon represents the 95% confidence interval."
  ) +
  scale_y_continuous(labels = label_percent()) +
  theme_minimal()
```

## 3. Visualizing ROC curve

An ROC (Receiver Operating Characteristic) curve shows how well a logistic model distinguishes between two classes. We use the yardstick package for a tidy workflow.

```{r}
#| fig-cap: "Figure 3: ROC Curve for the transmission prediction model. The area under the curve (AUC) indicates high predictive accuracy."

# 1. Get predictions from the original data
m_log_preds <- augment(m_log, type.predict = "response") |> 
  mutate(am = as.factor(am))

# 2. Calculate the ROC curve coordinates
roc_data <- roc_curve(m_log_preds, am, .fitted, event_level = "second")

# 3. Plot
autoplot(roc_data) +
  labs(
    title = "Model Performance: ROC Curve",
    subtitle = "Predicting Manual vs. Automatic Transmission",
    caption = "A diagonal line represents a model no better than random chance."
  ) +
  theme_minimal()
```

## 4. Captions and Figure Numbering

A plot without a proper caption is like a news story without a lead. Captions provide context, define abbreviations, and state the "takeaway" so the reader doesn't have to guess.

### Good vs. Bad Captions

| Type | Example | Why? |
|:-----------------------|:-----------------------|:-----------------------|
| **Bad** | "Plot of regression results." | It is redundant (the reader can see it's a plot) and provides no context. |
| **Good** | "Figure 1: Estimated change in MPG per unit increase in predictor. Error bars represent 95% CIs. Note that Weight and Horsepower show significant negative correlations with efficiency." | It defines the error bars, identifies the data source, and highlights the key finding. |

### How to add it in Quarto/ggplot

You can add two types of captions:

1.  **Inside `labs()` (as seen above):** Best for general notes and data sourcing.

2.  **In the Quarto Chunk Header:** Best for formal figure numbering in reports. **Using this you can cross-reference the figure in other parts of the text!**

#### Cross-Referencing (The Real Magic)

To enable automatic numbering (e.g., "Figure 1"), you must provide **both a label and a caption** in the code chunk header.

```{r}
#| fig-width: 4
#| fig-height: 3
#| label: fig-mpg-results
#| fig-cap: "Impact of Car Characteristics on Fuel Efficiency. Error bars represent 95% Confidence Intervals."
#| echo: fenced

ggplot(results_tidy, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high), width = 0.2) +
  geom_point(size = 3, color = "steelblue") +
  theme_minimal()
```

#### Why do we need both?

|  |  |  |
|------------------------|------------------------|------------------------|
| **Component** | **Example** | **Purpose** |
| **`label`** | `fig-mpg-results` | The **internal ID**. It must start with `fig-`. It enables the numbering engine and allows you to "call" this figure in your text. |
| **`fig-cap`** | `"Impact of Car..."` | The **external text**. This is the human-readable description that appears directly under the plot in your report. |

The primary reason Quarto uses this system is so you don't have to remember the numbers yourself. Once you have a label, you can reference it in your prose:

> As seen in `@fig-mpg-results`, weight is the most significant predictor...

Becomes

> As seen in @fig-mpg-results, weight is the most significant predictor...

The magic is that if you add a new plot above it later, Quarto will automatically re-index it to "Figure 4" without you changing a single word of text.

## Summary of Best Practices

1.  **Forest Plots for Coefficients**: Use `tidy(conf.int = TRUE)` to extract estimates and confidence intervals. A forest plot is the gold standard for showing which variables are "significant" (do they cross the 0 line?) and the magnitude of their impact.

2.  **Intentional Axis Sorting**: Never settle for alphabetical order. Use `reorder(term, estimate)` to sort your variables by the size of their effect. This creates a "ladder" effect that makes the strongest predictors stand out immediately.

3.  **Probability Curves for Logistic Regression**: Raw log-odds are nearly impossible to visualize. Use `plogis()` to transform model predictions into **probabilities** ($0$ to $100\%$). This allows students to see the "S-curve" and understand how the likelihood of an event changes across a range of values.

4.  **Model Performance with ROC**: Beyond individual predictors, use the ROC curve to visualize the overall accuracy of a classification model. A curve hugging the top-left corner represents a strong model, while the diagonal line represents random guessing.

5.  **Interpretive Captions**: A good caption is self-contained. Instead of saying "Plot of model," use the caption to state the finding (e.g., "Figure 1: Weight has the strongest negative impact on MPG...").

6.  **The Label + Caption Rule**: To enable automatic figure numbering in Quarto, every code chunk must include both `#| label: fig-xxx` and `#| fig-cap: "..."`.

7.  **Automated Cross-Referencing**: Use `@fig-label` in your text to create clickable links. Quarto will handle the numbering (Figure 1, Figure 2, etc.) for you, even if you move your plots around later.
